---
title: "Healthcare EDA"
author: “Kristin Lloyd”
format: 
  html:
    embed-resources: true
    code-fold: true
---

```{r}

library(readr)
df <- read_csv("data/clean_data/merged_data.csv")

head(df)

```

```{r}

options(repos = c(CRAN = "https://cran.r-project.org"))
install.packages("dplyr")
library(dplyr)

```

```{r}

head(df)
colnames(df) 

```

```{r}

library(ggplot2)
library(tidyr)
library(dplyr)

```

```{r}

# wide to long format

df_racial_mortality <- pivot_longer(
  data = df,
  cols = c(
    maternal_mortality_mothers_race_is_black_2017_2022,
    maternal_mortality_mothers_race_is_white_2017_2022
  ),
  names_to = "Race",
  values_to = "MortalityRate"
)

df_racial_mortality$Race <- ifelse(
  df_racial_mortality$Race == "maternal_mortality_mothers_race_is_black_2017_2022", 
  "Black",
  ifelse(
    df_racial_mortality$Race == "maternal_mortality_mothers_race_is_white_2017_2022",
    "White",
    df_racial_mortality$Race
  )
)

```

```{r}

# box plot for maternal mortality by race

ggplot(data = df_racial_mortality, 
       aes(x = Race, y = MortalityRate, fill = Race)) +
  geom_boxplot(alpha = 0.7) +
  labs(
    title = "Distribution of Maternal Mortality Rates by Race (2017-2022)",
    x = "Race",
    y = "Mortality Rate per 100,000 Live Births",
    caption = "Data source: Merged healthcare dataset"
  ) +
  scale_fill_manual(
    values = c("Black" = "#1f77b4", "White" = "#ff7f0e")
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    axis.title = element_text(size = 12),
    legend.position = "none"
  )

```

###

Null hypothesis: The maternal mortality rate for Black mothers is less than or equal to the rate for White mothers.

Alternate hypothesis: The maternal mortality rate for Black mothers is greater than the rate for White mothers

```{r}

black_rates <- na.omit(df$maternal_mortality_mothers_race_is_black_2017_2022)
white_rates <- na.omit(df$maternal_mortality_mothers_race_is_white_2017_2022)

t_test_result <- t.test(black_rates, white_rates, alternative = "greater")

print(t_test_result)

```

###

The p value is extremely small and less than .05. Therefore, we can reject the null hypothesis and say that black maternal mortality rates are greater than White maternal mortality rates. The lower bound, 4.91, says that we are 95% confident the true difference in means is greater than 4.91. 

```{r}

df_long_infant <- pivot_longer(
  data = df,
  cols = c(total_infant_mortality_white, 
           total_infant_mortality_black, 
           total_infant_mortality_hispanic),
  names_to = "Group",
  values_to = "InfantMortalityRate"
)

df_long_infant$Group <- ifelse(
  df_long_infant$Group == "total_infant_mortality_white", "White",
  ifelse(
    df_long_infant$Group == "total_infant_mortality_black", "Black",
    ifelse(
      df_long_infant$Group == "total_infant_mortality_hispanic", "Hispanic",
      df_long_infant$Group
    )
  )
)

```

```{r}

ggplot(data = df_long_infant, 
       aes(x = Group, y = InfantMortalityRate, fill = Group)) +
  geom_boxplot(alpha = 0.7) +
  labs(
    title = "Distribution of Infant Mortality Rates by Race/Ethnicity",
    x = "Race/Ethnicity",
    y = "Infant Mortality Rate per 1,000 Live Births",
    caption = "Data source: Merged healthcare dataset"
  ) +
  scale_fill_manual(
    values = c("White" = "#1f77b4", "Black" = "#ff7f0e", "Hispanic" = "#2ca02c")
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    axis.title = element_text(size = 12),
    legend.position = "none"
  )

```

###
Need to use a different statistical test for this since there are three groups with different variances. Or we can do 3 separate t-tests.

```{r}

mortality_plot <- pivot_longer(
  data = df,
  cols = starts_with("infant_mortality_"),
  names_to = "year",
  values_to = "mortality_rate"
)

mortality_plot$year <- as.numeric(substr(
  mortality_plot$year, 
  nchar("infant_mortality_") + 1, 
  nchar(mortality_plot$year)
))

mortality_plot <- mortality_plot[
  mortality_plot$abortion_policies == "most restrictive" & 
  !is.na(mortality_plot$abortion_policies),
]

ggplot(data = mortality_plot, 
       aes(x = year, y = mortality_rate,
           color = State, group = State)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  theme_minimal() +
  labs(
    title = "Infant Mortality Rates in Most Restrictive States (2017-2022)",
    x = "Year",
    y = "Infant Mortality Rate per 1,000 Live Births",
    color = "State",
    caption = "Data source: Merged healthcare dataset"
  ) +
  scale_x_continuous(breaks = 2017:2022) +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10)
  )

```

###
Some go up, some go down. Not very telling!

```{r}

# Add a new column for OB/GYN proportion in the original data frame
df <- df %>%
  filter(!is.na(abortion_policies) & 
         abortion_policies != "" & 
         abortion_policies != "NA") %>%
  mutate(
    abortion_policies = tolower(trimws(abortion_policies)),
    obgyn_proportion = (number_of_obstetricians_and_gynecologists_employed_2023 /
                        total_sumber_of_women_aged_15_44_2017) * 100000
  )

```

```{r}

create_ban_type_viz <- function(data, metric_col, title, y_label, plot_type = "bar") {
  ban_order <- c("total_abortion_ban",
                 "week_ban_6",
                 "week_ban_12",
                 "week_ban_15",
                 "week_ban_18_to_23",
                 "week_ban_24_to_26",
                 "no_gestational_ban")
  
  reshaped_data <- data %>%
    select(all_of(ban_order), !!sym(metric_col)) %>%
    pivot_longer(
      cols = all_of(ban_order),
      names_to = "ban_type",
      values_to = "has_ban"
    ) %>%
    filter(has_ban == 1) %>%
    group_by(ban_type) %>%
    summarise(avg_value = mean(!!sym(metric_col), na.rm = TRUE)) %>%
    mutate(
      ban_type = factor(ban_type, levels = ban_order),
      ban_numeric = as.numeric(factor(ban_type, levels = ban_order))
    )
  
  p <- ggplot(reshaped_data, aes(x = ban_type, y = avg_value))
  
  if (plot_type == "scatter") {
    model <- lm(avg_value ~ ban_numeric, data = reshaped_data)
    r_squared <- round(summary(model)$r.squared, 3)
    
    p <- p +
      geom_point(size = 4, color = "steelblue") +
      geom_smooth(aes(x = ban_numeric), method = "lm", se = TRUE, color = "red") +
      annotate("text", 
               x = length(ban_order) - 0.5,
               y = max(reshaped_data$avg_value),
               label = paste0("R² = ", r_squared),
               size = 4, hjust = 1)
  } else {

    p <- p + geom_bar(stat = "identity", fill = "steelblue")
  }
  
  p + labs(
    title = title,
    x = "Abortion Ban Type",
    y = y_label
  ) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
      axis.title = element_text(size = 12)
    )
}

```

```{r}

all_cause_mortality <- create_ban_type_viz(df, "all_cause_mortality_rate_per_100,000_women_ages 15_44_2022",
                                    "All Cause Mortality Rate Women 15-44, 2022",
                                    "Rate")

```

```{r}

print(all_cause_mortality)

```

###

The total abortion ban category is significantly higher despite all of the other categories being somewhat stable. 

# Maternal Mortality

```{r}

df_filtered <- df %>%
  filter(!is.na(`maternal_mortality_rates_2018_2022_per_100,000_live births`)) %>%
  filter(State != "US") %>%
  bind_rows(tibble(
    State = "US Average",
    `maternal_mortality_rates_2018_2022_per_100,000_live births` = 23.2,
    abortion_policies = "US Average"
  )) %>%
  mutate(
    abortion_policies = factor(
      abortion_policies,
      levels = c(
        "most restrictive", "very restrictive", "restrictive",
        "some restrictions/protections", 
        "protective", "very protective", "most protective", "US Average"
      )
    )
  )

ggplot(df_filtered, 
       aes(y = reorder(State, `maternal_mortality_rates_2018_2022_per_100,000_live births`), 
           x = `maternal_mortality_rates_2018_2022_per_100,000_live births`,
           fill = abortion_policies)) +

  geom_bar(stat = "identity", width = 0.7) +
  geom_text(aes(label = sprintf("%.1f", `maternal_mortality_rates_2018_2022_per_100,000_live births`)),
            hjust = -0.2,
            size = 3) +

  scale_fill_manual(values = c(
    "most restrictive" = "#bd0000",     
    "very restrictive" = "#ff0000",     
    "restrictive" = "#ff866b",    
    "some restrictions/protections" = "#ffe857", 
    "protective" = "#b6ff83",           
    "very protective" = "#2ae81e",    
    "most protective" = "#1e9400",    
    "US Average" = "#0056df"   
  )) +

  theme_minimal() +
  labs(
    title = "Maternal Mortality Rates (2018-2022) per 100,000 Live Births by State",
    x = "Maternal Mortality Rate per 100,000 Live Births",
    y = "",
    fill = "Abortion Policy"
  ) +

  theme(
    plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
    axis.title.x = element_text(size = 10),
    axis.text.y = element_text(size = 8, color = "grey30"),
    axis.text.x = element_text(size = 8),
    legend.position = "right",
    legend.title = element_text(size = 9),
    legend.text = element_text(size = 8),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA)
  ) +

  scale_x_continuous(
    limits = c(0, 45),
    breaks = seq(0, 40, by = 10),
    expand = expansion(mult = c(0, 0.1))
  )

```

```{r}

df_filtered <- df %>%
  filter(!is.na(`maternal_mortality_rates_2018_2022_per_100,000_live births`)) %>%
  filter(State != "US" & State != "Ohio" & State != "Nevada") %>%
  mutate(
    simplified_policies = case_when(
      abortion_policies %in% c("most restrictive", "very restrictive", "restrictive") ~ "Restrictive",
      abortion_policies %in% c("most protective", "very protective", "protective") ~ "Protective",
      TRUE ~ abortion_policies
    )
  ) %>%
  bind_rows(tibble(
    State = "US Average",
    `maternal_mortality_rates_2018_2022_per_100,000_live births` = 23.2,
    simplified_policies = "US Average"
  ))

ggplot(df_filtered, 
       aes(y = reorder(State, `maternal_mortality_rates_2018_2022_per_100,000_live births`), 
           x = `maternal_mortality_rates_2018_2022_per_100,000_live births`,
           fill = simplified_policies)) +
  geom_bar(stat = "identity", width = 0.7) +
  geom_text(aes(label = sprintf("%.1f", `maternal_mortality_rates_2018_2022_per_100,000_live births`)),
            hjust = -0.2,
            size = 3) +
  scale_fill_manual(values = c(
    "Restrictive" = "#ff0000",    
    "Protective" = "#4dbd00",    
    "US Average" = "#006eff"   
  )) +
  theme_minimal() +
  labs(
    title = "Maternal Mortality Rates (2018-2022) per 100,000 Live Births by State",
    x = "Maternal Mortality Rate per 100,000 Live Births",
    y = "",
    fill = "Abortion Policy"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
    axis.title.x = element_text(size = 10),
    axis.text.y = element_text(size = 8, color = "grey30"),
    axis.text.x = element_text(size = 8),
    legend.position = "right",
    legend.title = element_text(size = 9),
    legend.text = element_text(size = 8),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA)
  ) +
  scale_x_continuous(
    limits = c(0, 45),
    breaks = seq(0, 40, by = 10),
    expand = expansion(mult = c(0, 0.1))
  )

```

###
Separated only by protective and restrictive.

```{r}

install.packages("usmap")
library(usmap)

library()
abortion_map_data <- df %>%
  select(State, `percent_increase_in_clinician_provided_abortions_since_2020`) %>%
  filter(!is.na(`percent_increase_in_clinician_provided_abortions_since_2020`)) %>%
  mutate(
    state = State,
    change = `percent_increase_in_clinician_provided_abortions_since_2020`
  ) 

print("First few rows of our data:")
head(abortion_map_data)

plot_usmap(data = abortion_map_data, values = "change") +
  scale_fill_gradient2(
    name = "% Change",
    low = "#ff6b6b",   
    mid = "#ffffff",    
    high = "#4dbd00",  
    midpoint = 0,
    na.value = "gray80"  
  ) +
  theme(
    legend.position = "right",
    plot.title = element_text(size = 14, hjust = 0.5, face = "bold")
  ) +
  labs(title = "Change in Clinician-Provided Abortions Since 2020 by State")

```

###
Cool, you can see it goes up near the abortion ban states.

```{r}

library(corrplot)

healthcare_metrics <- df %>%
  select(
    overall_health_ranking,
    coverage_access_and_affordability,
    health_care_quality_and_prevention,
    health_outcomes,
    women_18_44_who_went_without_care_because_of_cost.1
  ) %>%
  rename(
    "Overall Health Ranking" = overall_health_ranking,
    "Coverage & Access" = coverage_access_and_affordability,
    "Healthcare Quality" = health_care_quality_and_prevention,
    "Health Outcomes" = health_outcomes,
    "Care Cost Barriers" = women_18_44_who_went_without_care_because_of_cost.1
  )

cor_matrix <- cor(healthcare_metrics, use = "complete.obs")

corrplot(cor_matrix, 
         method = "color",        
         type = "upper",           
         tl.col = "black",        
         tl.srt = 45,            
         addCoef.col = "black",    
         number.cex = 0.7,       
         col = colorRampPalette(c("#E41A1C", "#FFFFFF", "#4DAF4A"))(200),  
         tl.cex = 0.7,            
         cl.cex = 0.7,            
         diag = FALSE,           
         mar = c(0,0,1,0),        
         title = "Healthcare Metrics Correlations",
         family = "Arial",
         bg = "white",
         addgrid.col = "gray90")  

title("Correlations Between Healthcare Metrics", 
      line = -0.5, 
      cex.main = 1.2, 
      font.main = 2)

```

###
Love this

```{r}

library(corrplot)

healthcare_metrics <- df %>%
  select(
    overall_health_ranking,
    coverage_access_and_affordability,
    health_care_quality_and_prevention,
    health_outcomes,
    women_18_44_who_went_without_care_because_of_cost.1,
    `percent_uninsured_women_ages 19_64`,
    `teen_birth_rate_per_1,000_population_ages_15_19_2022.1`,
    `maternal_mortality_rates_2018_2022_per_100,000_live births`
  ) %>%
  rename(
    "Overall Health Ranking" = overall_health_ranking,
    "Coverage & Access" = coverage_access_and_affordability,
    "Healthcare Quality" = health_care_quality_and_prevention,
    "Health Outcomes" = health_outcomes,
    "Care Cost Barriers" = women_18_44_who_went_without_care_because_of_cost.1,
    "Uninsured Women %" = `percent_uninsured_women_ages 19_64`,
    "Teen Birth Rate" = `teen_birth_rate_per_1,000_population_ages_15_19_2022.1`,
    "Maternal Mortality" = `maternal_mortality_rates_2018_2022_per_100,000_live births`
  )

cor_matrix <- cor(healthcare_metrics, use = "complete.obs")

corrplot(cor_matrix, 
         method = "color",        
         type = "upper",           
         tl.col = "black",        
         tl.srt = 45,            
         addCoef.col = "black",    
         number.cex = 0.7,       
         col = colorRampPalette(c("#E41A1C", "#FFFFFF", "#4DAF4A"))(200),  
         tl.cex = 0.7,            
         cl.cex = 0.7,            
         diag = FALSE,           
         mar = c(0,0,1,0),        
         title = "Healthcare Metrics Correlations",
         family = "Arial",
         bg = "white",
         addgrid.col = "gray90")  

title("Correlations Between Healthcare Metrics", 
      line = -0.5, 
      cex.main = 1.2, 
      font.main = 2)

```

###
Can add more, have to decide which.