---
title: "Healthcare EDA"
author: “Kristin Lloyd”
format: 
  html:
    embed-resources: true
    code-fold: true
---

```{r}

library(readr)
df <- read_csv("../data/clean_data/merged_data.csv")

head(df)

```

```{r}

options(repos = c(CRAN = "https://cran.r-project.org"))
install.packages("dplyr")
library(dplyr)

```

```{r}

head(df)
colnames(df) 

```

```{r}

library(ggplot2)
library(tidyr)
library(dplyr)

```

```{r}

# wide to long format

df_racial_mortality <- pivot_longer(
  data = df,
  cols = c(
    maternal_mortality_mothers_race_is_black_2017_2022,
    maternal_mortality_mothers_race_is_white_2017_2022
  ),
  names_to = "Race",
  values_to = "MortalityRate"
)

df_racial_mortality$Race <- ifelse(
  df_racial_mortality$Race == "maternal_mortality_mothers_race_is_black_2017_2022", 
  "Black",
  ifelse(
    df_racial_mortality$Race == "maternal_mortality_mothers_race_is_white_2017_2022",
    "White",
    df_racial_mortality$Race
  )
)

```

```{r}

# box plot for maternal mortality by race

ggplot(data = df_racial_mortality, 
       aes(x = Race, y = MortalityRate, fill = Race)) +
  geom_boxplot(alpha = 0.7) +
  labs(
    title = "Distribution of Maternal Mortality Rates by Race (2017-2022)",
    x = "Race",
    y = "Mortality Rate per 100,000 Live Births",
    caption = "Data source: Merged healthcare dataset"
  ) +
  scale_fill_manual(
    values = c("Black" = "#1f77b4", "White" = "#ff7f0e")
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    axis.title = element_text(size = 12),
    legend.position = "none"
  )

```

###
Null hypothesis: The maternal mortality rate for Black mothers is less than or equal to the rate for White mothers.

Alternate hypothesis: The maternal mortality rate for Black mothers is greater than the rate for White mothers

```{r}

black_rates <- na.omit(df$maternal_mortality_mothers_race_is_black_2017_2022)
white_rates <- na.omit(df$maternal_mortality_mothers_race_is_white_2017_2022)

t_test_result <- t.test(black_rates, white_rates, alternative = "greater")

print(t_test_result)

```

###
The p value is extremely small and less than .05. Therefore, we can reject the null hypothesis and say that black maternal mortality rates are greater than White maternal mortality rates. The lower bound, 4.91, says that we are 95% confident the true difference in means is greater than 4.91. 

```{r}

df_long_infant <- pivot_longer(
  data = df,
  cols = c(total_infant_mortality_white, 
           total_infant_mortality_black, 
           total_infant_mortality_hispanic),
  names_to = "Group",
  values_to = "InfantMortalityRate"
)

df_long_infant$Group <- ifelse(
  df_long_infant$Group == "total_infant_mortality_white", "White",
  ifelse(
    df_long_infant$Group == "total_infant_mortality_black", "Black",
    ifelse(
      df_long_infant$Group == "total_infant_mortality_hispanic", "Hispanic",
      df_long_infant$Group
    )
  )
)

```

```{r}

ggplot(data = df_long_infant, 
       aes(x = Group, y = InfantMortalityRate, fill = Group)) +
  geom_boxplot(alpha = 0.7) +
  labs(
    title = "Distribution of Infant Mortality Rates by Race/Ethnicity",
    x = "Race/Ethnicity",
    y = "Infant Mortality Rate per 1,000 Live Births",
    caption = "Data source: Merged healthcare dataset"
  ) +
  scale_fill_manual(
    values = c("White" = "#1f77b4", "Black" = "#ff7f0e", "Hispanic" = "#2ca02c")
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    axis.title = element_text(size = 12),
    legend.position = "none"
  )

```

###
Need to use a different statistical test for this since there are three groups with different variances. Or we can do 3 separate t-tests.

```{r}

mortality_plot <- pivot_longer(
  data = df,
  cols = starts_with("infant_mortality_"),
  names_to = "year",
  values_to = "mortality_rate"
)

mortality_plot$year <- as.numeric(substr(
  mortality_plot$year, 
  nchar("infant_mortality_") + 1, 
  nchar(mortality_plot$year)
))

mortality_plot <- mortality_plot[
  mortality_plot$abortion_policies == "most restrictive" & 
  !is.na(mortality_plot$abortion_policies),
]

ggplot(data = mortality_plot, 
       aes(x = year, y = mortality_rate,
           color = State, group = State)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  theme_minimal() +
  labs(
    title = "Infant Mortality Rates in Most Restrictive States (2017-2022)",
    x = "Year",
    y = "Infant Mortality Rate per 1,000 Live Births",
    color = "State",
    caption = "Data source: Merged healthcare dataset"
  ) +
  scale_x_continuous(breaks = 2017:2022) +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10)
  )

```

###
Some go up, some go down. Not very telling!

```{r}

policy_order <- c("most protective", "very protective", "protective", 
                 "some restrictions/protections", "restrictive", 
                 "very restrictive", "most restrictive")

```

```{r}

create_policy_viz <- function(data, metric_col, title, y_label, plot_type = "bar") {
  processed_data <- data %>%
    filter(!is.na(abortion_policies) & 
           abortion_policies != "" & 
           abortion_policies != "NA") %>%
    mutate(abortion_policies = tolower(trimws(abortion_policies))) %>%
    group_by(abortion_policies) %>%
    summarise(avg_value = mean(!!sym(metric_col), na.rm = TRUE)) %>%
    filter(abortion_policies %in% policy_order) %>%
    mutate(
      abortion_policies = factor(abortion_policies, levels = policy_order),
      policy_numeric = as.numeric(factor(abortion_policies, levels = policy_order))
    )
  
  p <- ggplot(processed_data, aes(x = abortion_policies, y = avg_value))
  
  if(plot_type == "scatter") {
    model <- lm(avg_value ~ policy_numeric, data = processed_data)
    r_squared <- round(summary(model)$r.squared, 3)
    
    p <- p +
      geom_point(size = 4, color = "steelblue") +
      geom_smooth(aes(x = policy_numeric), method = "lm", se = TRUE, color = "red") +
      annotate("text", 
              x = length(policy_order) - 0.5,
              y = max(processed_data$avg_value),
              label = paste0("R² = ", r_squared),
              size = 4, hjust = 1)
  } else {
    p <- p + geom_bar(stat = "identity", fill = "steelblue")
  }
  
  p + labs(
    title = title,
    x = "Abortion Policy",
    y = y_label
  ) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
      axis.title = element_text(size = 12)
    )
}

```

```{r}

health_ranking <- create_policy_viz(df,
    "overall_health_ranking",
    "Overall Health Ranking",
    "Ranking", 
    "scatter")

```

```{r}

print(health_ranking)

```

###
R-squared is .917, showing strong linear correlation. Although this is the average healthcare rankings of the states, very interesting. Should explore this more with other plots and maybe some sort of test.

```{r}

coverage_access_affordability <- create_policy_viz(df,
    "coverage_access_and_affordability",
    "Coverage, Access, and Affordability",
    "Ranking", 
    "scatter")

```

```{r}

print(coverage_access_affordability)

```

###
.826, also a strong correlation. Should explore this one more.

```{r}

healthcare_quality_prevention <- create_policy_viz(df,
    "health_care_quality_and_prevention",
    "Healthcare Quality and Prevention",
    "Ranking", 
    "scatter")

```

```{r}

print(healthcare_quality_prevention)

```

###
.618, less strong but would also like to explore further. 

```{r}

health_outcomes <- create_policy_viz(df,
    "health_outcomes",
    "Health Outcomes",
    "Ranking", 
    "scatter")

```

```{r}

print(health_outcomes)

```

###
Strong linear correlation again!

```{r}

healthcare_cost <- create_policy_viz(df,
   "women_18_44_who_went_without_care_because_of_cost.1",
   "Women Without Care Due to Cost",
   "Percentage", 
   "scatter")

```

```{r}

print(healthcare_cost)

```

###
Very strong correlation.

```{r}

uninsured_women <- create_policy_viz(df,
   "percent_uninsured_women_ages 19_64",
   "Uninsured Women Ages 19-64",
   "Percentage", 
   "scatter")

```

```{r}

print(uninsured_women)

```

###
Again, very strong correlation. 

```{r}

uninsured_pregnancy <- create_policy_viz(df,
   "percent_uninsured_women_during_pregnancy_2021",
   "Uninsured During Pregnancy",
   "Percentage", 
   "scatter")

```

```{r}

print(uninsured_pregnancy)

```

###
.4, not a great fit. What else can I do with this?

```{r}

teen_birth <- create_policy_viz(df,
   "teen_birth_rate_per_1,000_population_ages_15_19_2022.1",
   "Teen Birth Rate",
   "Rate per 1,000", 
   "scatter")

```

```{r}

print(teen_birth)

```

###
Very strong, teens give birth in restrictive states more.This could effect their lives and education.

```{r}

all_birth <- create_policy_viz(df,
   "birth_rate_per_1,000_women_ages_15_44",
   "Birth Rate",
   "Rate per 1,000", 
   "scatter")

```

```{r}

print(all_birth)

```

###
.73, strong-ish correlation. I wonder if anything is significant about this.

```{r}

workforce <- create_policy_viz(df,
   "maternity_care_workforce_per_100,000_women_ages_15_44",
   "Maternity Care Workforce",
   "Per 100,000 Women", 
   "scatter")

```

```{r}

print(workforce)

```

###
Negative correlation. This means that mothers work less in restrictive states. Why?

```{r}

no_care <- create_policy_viz(df,
   "percent_women_ages_18_44_without_a_usual_source_of_care",
   "Women Without Usual Source of Care",
   "Percentage", 
   "scatter")

```

```{r}

print(no_care)

```

```{r}

medicaid_births <- create_policy_viz(df,
   "percent_of_births_financed_by_medicaid",
   "Births Financed by Medicaid",
   "Percentage", 
   "scatter")

```

```{r}

print(medicaid_births)

```

```{r}

depression <- create_policy_viz(df,
   "women_with_depression_before_or_during_pregnancy_2021",
   "Depression Before/During Pregnancy",
   "Percentage", 
   "scatter")

```

```{r}

print(depression)

```

```{r}

postpartum_depression <- create_policy_viz(df,
   "percent_women_with_postpartum_depression_2021",
   "Postpartum Depression",
   "Percentage", 
   "scatter")

```

```{r}

print(postpartum_depression)

```

```{r}

depression_screening <- create_policy_viz(df,
   "percent_women_who_received_a_postpartum_depression_screening_2021",
   "Postpartum Depression Screening",
   "Percentage", 
   "scatter")

```

```{r}

print(depression_screening)

```

```{r}

violence <- create_policy_viz(df,
   "percent_women_experiencing_intimate_partner_violence_2021",
   "Intimate Partner Violence",
   "Percentage", 
   "scatter")

```

```{r}

print(violence)

```

```{r}

increase_in_abortions <- create_policy_viz(df,
   "percent_increase_in_clinician_provided_abortions_since_2020",
   "Increase in Clinician-Provided Abortions",
   "Percentage Change Since 2020", 
   "scatter")

```

```{r}

print(increase_in_abortions)

```

```{r}

maternal_mortality <- create_policy_viz(df,
   "maternal_mortality_rates_2018_2022_per_100,000_live births",
   "Maternal Mortality Rate, 2018-2022",
   "Rate",
   "scatter")

```

```{r}

print(maternal_mortality)

```

```{r}

obgyn_proportion_grouped <- df %>%
 filter(!is.na(abortion_policies) &
        abortion_policies != "" &
        abortion_policies != "NA") %>%
 mutate(
   abortion_policies = tolower(trimws(abortion_policies)),
   obgyn_proportion = (number_of_obstetricians_and_gynecologists_employed_2023 /
                      total_sumber_of_women_aged_15_44_2017) * 100000,
   policy_group = case_when(
     abortion_policies %in% c("most protective", "very protective", "protective") ~ "Protective",
     abortion_policies %in% c("restrictive", "very restrictive", "most restrictive") ~ "Restrictive",
     TRUE ~ "Mixed"
   )
 ) %>%
 filter(policy_group != "Mixed")  

ggplot(obgyn_proportion_grouped, 
      aes(x = policy_group, y = obgyn_proportion, fill = policy_group)) +
 geom_boxplot() +
 labs(
   title = "OB/GYNs per 100,000 Women (Ages 15-44) by Policy Group",
   subtitle = "2023 OB/GYN Employment Data, 2017 Population Data",
   x = "Policy Group",
   y = "OB/GYNs per 100,000 Women Ages 15-44"
 ) +
 scale_fill_manual(values = c("Protective" = "#1f77b4", "Restrictive" = "#ff7f0e")) +
 theme_minimal() +
 theme(
   plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
   plot.subtitle = element_text(hjust = 0.5, size = 10),
   axis.title = element_text(size = 12),
   legend.position = "none")

```

###

Null hypothesis: The mean OB/GYN proportion for the "Protective" and "Restrictive" policy groups is equal. 

Alternate hypothesis: The mean OB/GYN proportion for the "Protective" and "Restrictive" policy groups is not equal.


```{r}

t_test_result <- t.test(
  obgyn_proportion ~ policy_group,
  data = obgyn_proportion_grouped,
  subset = policy_group %in% c("Protective", "Restrictive")
)

print(t_test_result)

```

###

The small p-value (0.009633) suggests rejecting the null hypothesis at a 5% significance level, indicating a significant difference in OB/GYN proportions between the two groups.


```{r}

create_ban_type_viz <- function(data, metric_col, title, y_label, plot_type = "bar") {
  ban_order <- c("total_abortion_ban",
                 "week_ban_6",
                 "week_ban_12",
                 "week_ban_15",
                 "week_ban_18_to_23",
                 "week_ban_24_to_26",
                 "no_gestational_ban")
  
  reshaped_data <- data %>%
    select(all_of(ban_order), !!sym(metric_col)) %>%
    pivot_longer(
      cols = all_of(ban_order),
      names_to = "ban_type",
      values_to = "has_ban"
    ) %>%
    filter(has_ban == 1) %>%
    group_by(ban_type) %>%
    summarise(avg_value = mean(!!sym(metric_col), na.rm = TRUE)) %>%
    mutate(
      ban_type = factor(ban_type, levels = ban_order),
      ban_numeric = as.numeric(factor(ban_type, levels = ban_order))
    )
  
  p <- ggplot(reshaped_data, aes(x = ban_type, y = avg_value))
  
  if (plot_type == "scatter") {
    model <- lm(avg_value ~ ban_numeric, data = reshaped_data)
    r_squared <- round(summary(model)$r.squared, 3)
    
    p <- p +
      geom_point(size = 4, color = "steelblue") +
      geom_smooth(aes(x = ban_numeric), method = "lm", se = TRUE, color = "red") +
      annotate("text", 
               x = length(ban_order) - 0.5,
               y = max(reshaped_data$avg_value),
               label = paste0("R² = ", r_squared),
               size = 4, hjust = 1)
  } else {

    p <- p + geom_bar(stat = "identity", fill = "steelblue")
  }
  
  p + labs(
    title = title,
    x = "Abortion Ban Type",
    y = y_label
  ) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
      axis.title = element_text(size = 12)
    )
}

```

```{r}

all_cause_mortality <- create_ban_type_viz(df, "all_cause_mortality_rate_per_100,000_women_ages 15_44_2022",
                                    "All Cause Mortality Rate Women 15-44, 2022",
                                    "Rate")

```

```{r}

print(all_cause_mortality)

```

###

The total abortion ban category is significantly higher despite all of the other categories being somewhat stable. 

```{r}

df_filtered <- df %>%
  filter(!is.na(`maternal_mortality_rates_2018_2022_per_100,000_live births`)) %>%
  filter(State != "US") %>%
  bind_rows(tibble(
    State = "US Average",
    `maternal_mortality_rates_2018_2022_per_100,000_live births` = 23.2,
    abortion_policies = "US Average"
  )) %>%
  mutate(
    abortion_policies = factor(
      abortion_policies,
      levels = c(
        "most restrictive", "very restrictive", "restrictive",
        "some restrictions/protections", 
        "protective", "very protective", "most protective", "US Average"
      )
    )
  )

ggplot(df_filtered, 
       aes(y = reorder(State, `maternal_mortality_rates_2018_2022_per_100,000_live births`), 
           x = `maternal_mortality_rates_2018_2022_per_100,000_live births`,
           fill = abortion_policies)) +

  geom_bar(stat = "identity", width = 0.7) +
  geom_text(aes(label = sprintf("%.1f", `maternal_mortality_rates_2018_2022_per_100,000_live births`)),
            hjust = -0.2,
            size = 3) +

  scale_fill_manual(values = c(
    "most restrictive" = "#bd0000",     
    "very restrictive" = "#ff0000",     
    "restrictive" = "#ff866b",    
    "some restrictions/protections" = "#ffe857", 
    "protective" = "#b6ff83",           
    "very protective" = "#2ae81e",    
    "most protective" = "#74ff6b",    
    "US Average" = "#4dbd00"   
  )) +

  theme_minimal() +
  labs(
    title = "Maternal Mortality Rates (2018-2022) per 100,000 Live Births by State",
    x = "Maternal Mortality Rate per 100,000 Live Births",
    y = "",
    fill = "Abortion Policy"
  ) +

  theme(
    plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
    axis.title.x = element_text(size = 10),
    axis.text.y = element_text(size = 8, color = "grey30"),
    axis.text.x = element_text(size = 8),
    legend.position = "right",
    legend.title = element_text(size = 9),
    legend.text = element_text(size = 8),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA)
  ) +

  scale_x_continuous(
    limits = c(0, 45),
    breaks = seq(0, 40, by = 10),
    expand = expansion(mult = c(0, 0.1))
  )

```
  
```{r}

df_filtered <- df %>%
  filter(!is.na(`maternal_mortality_rates_2018_2022_per_100,000_live births`)) %>%
  filter(State != "US" & State != "Ohio" & State != "Nevada") %>%
  bind_rows(tibble(
    State = "US Average",
    `maternal_mortality_rates_2018_2022_per_100,000_live births` = 23.2,
    abortion_policies = "US Average"
  ))

# Create the plot
ggplot(df_filtered, 
       aes(y = reorder(State, `maternal_mortality_rates_2018_2022_per_100,000_live births`), 
           x = `maternal_mortality_rates_2018_2022_per_100,000_live births`,
           fill = ifelse(State == "US Average", "US Average", abortion_policies))) +
  geom_bar(stat = "identity", width = 0.7) +
  geom_text(aes(label = sprintf("%.1f", `maternal_mortality_rates_2018_2022_per_100,000_live births`)),
            hjust = -0.2,
            size = 3) +
  scale_fill_manual(values = c(
    "most restrictive" = "#ff0000",     
    "very restrictive" = "#ff0000",     
    "restrictive" = "#ff0000",    
    "some restrictions/protections" = "#ffe800", 
    "protective" = "#4dbd00",           
    "very protective" = "#4dbd00",    
    "most protective" = "#4dbd00",    
    "US Average" = "#006eff"   
  )) +
  theme_minimal() +
  labs(
    title = "Maternal Mortality Rates (2018-2022) per 100,000 Live Births by State",
    x = "Maternal Mortality Rate per 100,000 Live Births",
    y = "",
    fill = "Abortion Policy"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
    axis.title.x = element_text(size = 10),
    axis.text.y = element_text(size = 8, color = "grey30"),
    axis.text.x = element_text(size = 8),
    legend.position = "right",
    legend.title = element_text(size = 9),
    legend.text = element_text(size = 8),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA)
  ) +
  scale_x_continuous(
    limits = c(0, 45),
    breaks = seq(0, 40, by = 10),
    expand = expansion(mult = c(0, 0.1))
  )

```

```{r}

df_filtered <- df %>%
  filter(!is.na(`maternal_mortality_rates_2018_2022_per_100,000_live births`)) %>%
  filter(State != "US" & State != "Ohio" & State != "Nevada") %>%
  mutate(
    simplified_policies = case_when(
      abortion_policies %in% c("most restrictive", "very restrictive", "restrictive") ~ "Restrictive",
      abortion_policies %in% c("most protective", "very protective", "protective") ~ "Protective",
      TRUE ~ abortion_policies
    )
  ) %>%
  bind_rows(tibble(
    State = "US Average",
    `maternal_mortality_rates_2018_2022_per_100,000_live births` = 23.2,
    simplified_policies = "US Average"
  ))

ggplot(df_filtered, 
       aes(y = reorder(State, `maternal_mortality_rates_2018_2022_per_100,000_live births`), 
           x = `maternal_mortality_rates_2018_2022_per_100,000_live births`,
           fill = simplified_policies)) +
  geom_bar(stat = "identity", width = 0.7) +
  geom_text(aes(label = sprintf("%.1f", `maternal_mortality_rates_2018_2022_per_100,000_live births`)),
            hjust = -0.2,
            size = 3) +
  scale_fill_manual(values = c(
    "Restrictive" = "#ff0000",    
    "Protective" = "#4dbd00",    
    "US Average" = "#006eff"   
  )) +
  theme_minimal() +
  labs(
    title = "Maternal Mortality Rates (2018-2022) per 100,000 Live Births by State",
    x = "Maternal Mortality Rate per 100,000 Live Births",
    y = "",
    fill = "Abortion Policy"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
    axis.title.x = element_text(size = 10),
    axis.text.y = element_text(size = 8, color = "grey30"),
    axis.text.x = element_text(size = 8),
    legend.position = "right",
    legend.title = element_text(size = 9),
    legend.text = element_text(size = 8),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA)
  ) +
  scale_x_continuous(
    limits = c(0, 45),
    breaks = seq(0, 40, by = 10),
    expand = expansion(mult = c(0, 0.1))
  )

```

###

Can I do some sort of ANOVA that does not require equal amounts in each group?

###

Missing visualization on this

percent_increase_in_clinician_provided_abortions_since_2020


