---
title: "Healthcare Linear Regression"
author: “Kristin Lloyd”
format: 
  html:
    embed-resources: true
    code-fold: true
---

```{r}

library(readr)
df <- read_csv("data/clean_data/merged_data.csv")

head(df)

```

```{r}

options(repos = c(CRAN = "https://cran.r-project.org"))
install.packages("dplyr")
library(dplyr)

```

```{r}

library(ggplot2)
library(tidyr)
library(dplyr)

```

# Linear Regression

```{r}

policy_order <- c("most protective", "very protective", "protective", 
                 "some restrictions/protections", "restrictive", 
                 "very restrictive", "most restrictive")

```

```{r}

create_policy_viz <- function(data, metric_col, title, y_label, plot_type = "bar") {
  processed_data <- data %>%
    filter(!is.na(abortion_policies) & 
           abortion_policies != "" & 
           abortion_policies != "NA") %>%
    mutate(abortion_policies = tolower(trimws(abortion_policies))) %>%
    group_by(abortion_policies) %>%
    summarise(avg_value = mean(!!sym(metric_col), na.rm = TRUE)) %>%
    filter(abortion_policies %in% policy_order) %>%
    mutate(
      abortion_policies = factor(abortion_policies, levels = policy_order),
      policy_numeric = as.numeric(factor(abortion_policies, levels = policy_order))
    )
  
  p <- ggplot(processed_data, aes(x = abortion_policies, y = avg_value))
  
  if(plot_type == "scatter") {
    model <- lm(avg_value ~ policy_numeric, data = processed_data)
    r_squared <- round(summary(model)$r.squared, 3)
    
    p <- p +
      geom_point(size = 4, color = "steelblue") +
      geom_smooth(aes(x = policy_numeric), method = "lm", se = TRUE, color = "red") +
      annotate("text", 
              x = length(policy_order) - 0.5,
              y = max(processed_data$avg_value),
              label = paste0("R² = ", r_squared),
              size = 4, hjust = 1)
  } else {
    p <- p + geom_bar(stat = "identity", fill = "steelblue")
  }
  
  p + labs(
    title = title,
    x = "Abortion Policy",
    y = y_label
  ) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
      axis.title = element_text(size = 12)
    )
}

```

## Health Ranking 

```{r}

health_ranking <- create_policy_viz(df,
    "overall_health_ranking",
    "Overall Health Ranking",
    "Ranking", 
    "scatter")

```

```{r}

print(health_ranking)

```

### Interpretation

R-squared is .917, showing strong linear correlation.

## Coverage, Access, Affordability

```{r}

coverage_access_affordability <- create_policy_viz(df,
    "coverage_access_and_affordability",
    "Coverage, Access, and Affordability",
    "Ranking", 
    "scatter")

```

```{r}

print(coverage_access_affordability)

```

### Interpretation

R-squared is .826, showing strong linear correlation.

## Healthcare Quality

```{r}

healthcare_quality_prevention <- create_policy_viz(df,
    "health_care_quality_and_prevention",
    "Healthcare Quality and Prevention",
    "Ranking", 
    "scatter")

```

```{r}

print(healthcare_quality_prevention)

```

### Interpretation

R-squared is .618, showing somewhat of a correlation. 

## Health Outcomes

```{r}

health_outcomes <- create_policy_viz(df,
    "health_outcomes",
    "Health Outcomes",
    "Ranking", 
    "scatter")

```

```{r}

print(health_outcomes)

```

### Interpretation

R-squared is .878, showing a strong linear correlation.

## Healthcare Cost

```{r}

healthcare_cost <- create_policy_viz(df,
   "women_18_44_who_went_without_care_because_of_cost.1",
   "Women Without Care Due to Cost",
   "Percentage", 
   "scatter")

```

```{r}

print(healthcare_cost)

```

### Interpretation

R-squared is .913, showing a strong linear correlation.

## Uninsured Women

```{r}

uninsured_women <- create_policy_viz(df,
   "percent_uninsured_women_ages 19_64",
   "Uninsured Women Ages 19-64",
   "Percentage", 
   "scatter")

```

```{r}

print(uninsured_women)

```

### Interpretation

R-squared is .893, showing a strong linear correlation.

## Uninsured during pregnancy

```{r}

uninsured_pregnancy <- create_policy_viz(df,
   "percent_uninsured_women_during_pregnancy_2021",
   "Uninsured During Pregnancy",
   "Percentage", 
   "scatter")

```

```{r}

print(uninsured_pregnancy)

```

### Interpretation

R-squared is .436, showing a poor fit. 

## Teen Birth

```{r}

teen_birth <- create_policy_viz(df,
   "teen_birth_rate_per_1,000_population_ages_15_19_2022.1",
   "Teen Birth Rate",
   "Rate per 1,000", 
   "scatter")

```

```{r}

print(teen_birth)

```

### Interpretation

R-squared is .8333, showing strong linear correlation. If teens are giving birth at higher rates in abortion ban states, they may not reach their full potential education wise. Further, these teenagers are probably not financially stable when giving birth.  

## All Birth

```{r}

all_birth <- create_policy_viz(df,
   "birth_rate_per_1,000_women_ages_15_44",
   "Birth Rate",
   "Rate per 1,000", 
   "scatter")

```

```{r}

print(all_birth)

```

### Interpretation

R-squared is .73, states with stricter abortion bans seem to have higher birth rates.

## Workforce

```{r}

workforce <- create_policy_viz(df,
   "maternity_care_workforce_per_100,000_women_ages_15_44",
   "Maternity Care Workforce",
   "Per 100,000 Women", 
   "scatter")

```

```{r}

print(workforce)

```

### Interpretation

R-squred is .872 and R is negative.

## Women who went without care due to cost

```{r}

no_care <- create_policy_viz(df,
   "percent_women_ages_18_44_without_a_usual_source_of_care",
   "Women Without Usual Source of Care",
   "Percentage", 
   "scatter")

```

```{r}

print(no_care)

```

### Interpretation

R-squared is .5

## Medicaid Births

```{r}

medicaid_births <- create_policy_viz(df,
   "percent_of_births_financed_by_medicaid",
   "Births Financed by Medicaid",
   "Percentage", 
   "scatter")

```

```{r}

print(medicaid_births)

```

### Interpretation

R-squared is 0.049, no correlation.

## Depression

```{r}

depression <- create_policy_viz(df,
   "women_with_depression_before_or_during_pregnancy_2021",
   "Depression Before/During Pregnancy",
   "Percentage", 
   "scatter")

```

```{r}

print(depression)

```

### Interpretation

R-squared is 0.13, depression is very common.

## Postpartum depression

```{r}

postpartum_depression <- create_policy_viz(df,
   "percent_women_with_postpartum_depression_2021",
   "Postpartum Depression",
   "Percentage", 
   "scatter")

```

```{r}

print(postpartum_depression)

```

### Interpretation

R-squared is .46

## Depression Screenings

```{r}

depression_screening <- create_policy_viz(df,
   "percent_women_who_received_a_postpartum_depression_screening_2021",
   "Postpartum Depression Screening",
   "Percentage", 
   "scatter")

```

```{r}

print(depression_screening)

```

### Interpretation

R-squared is .63 and negative, shows that women are given less screenings in abortion ban states. It is exceptionally low for most restrictive states. 

### Violence

```{r}

violence <- create_policy_viz(df,
   "percent_women_experiencing_intimate_partner_violence_2021",
   "Intimate Partner Violence",
   "Percentage", 
   "scatter")

```

```{r}

print(violence)

```

### Interpretation

R-squared is .163, no correlation except way higher for states where abortion is fully banned. 

## Maternal Mortality

```{r}

maternal_mortality <- create_policy_viz(df,
   "maternal_mortality_rates_2018_2022_per_100,000_live births",
   "Maternal Mortality Rate, 2018-2022",
   "Rate",
   "scatter")

```

```{r}

print(maternal_mortality)

```

### Interpretation

R-squared is .844, very high correlation

```{r}

bachelors_or_higher <- create_policy_viz(df,
   "twenty_five_and_older_had_bachelors_by_higher_2021",
   "% 25 and older with a Bachelors or Higher, 2021",
   "Rate",
   "scatter")

```

```{r}

print(bachelors_or_higher)

```

### Interpretation

R-squared is .779, strong negative correlation. Especially low for most restrictive,

```{r}

concraception <- create_policy_viz(df,
   "state_requires_coverage_of_prescription_contraception",
   "State Requires Coverage of Prescription Coverage",
   "Rate",
   "scatter")

```

```{r}

print(concraception)

```

### Interpretation

R-squared is .882, very strong negative correlation

## Proportion of OBGYN

```{r}

df$proportion_obgyns_to_women <- df$number_of_obstetricians_and_gynecologists_employed_2023 / 
                                   df$total_sumber_of_women_aged_15_44_2017

```

```{r}

obgyn <- create_policy_viz(df,
   "proportion_obgyns_to_women",
   "State Requires Coverage of Prescription Coverage",
   "Rate",
   "scatter")

```

```{r}

print(obgyn)

```

### Interpretation

R-squared is .531, which is weak correlation. While this is not a great correlation, we can see from the chartt that the difference between having protections vs. restrictions is very big. What plot would be better for this?