---
title: "Maternal Health EDA"
author: "Kristin Lloyd"
format: 
  html:
    embed-resources: true
    code-fold: true
editor: visual
---

```{r}

library(tidyverse)
df <- read_csv("../data/clean_data/merged_data.csv", show_col_types = FALSE)
df <- df[!is.na(df$abortion_policies), ]
df_test <- df[df$State != "US", ]
```

## Maternal Mortality, by Abortion Policy

```{r}

df$abortion_policies <- factor(df$abortion_policies, 
                              levels = c("most protective", "very protective", 
                                         "protective", "some restrictions/protections", 
                                         "restrictive", "very restrictive", "most restrictive"))
# Kruskal-Wallis test
kruskal.test(`maternal_mortality_rates_2018_2022_per_100,000_live births` ~ abortion_policies, data = df)

# If Kruskal-Wallis is significant, run pairwise Wilcoxon tests
pairwise.wilcox.test(df$`maternal_mortality_rates_2018_2022_per_100,000_live births`, 
                     df$abortion_policies,
                     p.adjust.method = "bonferroni")

# accompanying boxplot visualization
ggplot(df, aes(x = abortion_policies, y = `maternal_mortality_rates_2018_2022_per_100,000_live births`, fill=abortion_policies)) +
  geom_boxplot() +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 8, angle = 45, hjust = 1),
        axis.title.y = element_text(size = 10)) + 
  scale_fill_manual(values = c("#1c7416", "#68bb59", "#acdf87", "#fab733", "#ff6242", "#ff0000", "#c61a09")) +
  labs(x = "Abortion Policies", 
       y = "Maternal Morality rate per 100,000",
       title = "Maternal Mortality Rate by Abortion Policy") +
  theme(plot.title = element_text(size = 20, hjust = 0.5))

```

```{r}

df <- df[!is.na(df$`maternal_mortality_rates_2018_2022_per_100,000_live births`), ]

# Kruskal-Wallis test
kruskal.test(`maternal_mortality_rates_2018_2022_per_100,000_live births` ~ abortion_policies, data = df)

# If Kruskal-Wallis is significant, run pairwise Wilcoxon tests
pairwise.wilcox.test(df$`maternal_mortality_rates_2018_2022_per_100,000_live births`, 
                     df$abortion_policies,
                     p.adjust.method = "bonferroni")

# For visualization
# Reorder the factor levels
df$abortion_policies <- factor(df$abortion_policies, 
                              levels = c("most protective", "very protective", 
                                         "protective", "some restrictions/protections", 
                                         "restrictive", "very restrictive", "most restrictive"))

# Create the plot
# accompanying boxplot visualization
ggplot(df, aes(x = abortion_policies, y = `maternal_mortality_rates_2018_2022_per_100,000_live births`, fill=abortion_policies)) +
  geom_boxplot() +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 8, angle = 45, hjust = 1),
        axis.title.y = element_text(size = 10)) + 
  scale_fill_manual(values = c("#1c7416", "#68bb59", "#acdf87", "#fab733", "#ff6242", "#ff0000", "#c61a09")) +
  labs(x = "Abortion Policies", 
       y = "Maternal Morality rate per 100,000",
       title = "Maternal Mortality Rate by Abortion Policy") +
  theme(plot.title = element_text(size = 20, hjust = 0.5))
```

The Kruskal-Wallis test indicates a statistically significant difference in maternal mortality rates between at least two of the abortion policy groups, with p equaling 0.0097. This suggests that maternal mortality rates are not evenly distributed across the levels of abortion policies. A significant difference is observed between "most restrictive" and "protective" abortion policies (p = 0.0027).

```{r}

set.seed(5100)

df_test <- df[df$State != "US", ]

set.seed(5100)

permutation_test <- function(data, group_col, value_col, n_permutations = 100000) {
  observed_stat <- summary(aov(data[[value_col]] ~ data[[group_col]]))[[1]][["F value"]][1]

  perm_stats <- replicate(n_permutations, {
    shuffled_values <- sample(data[[value_col]])
    perm_stat <- summary(aov(shuffled_values ~ data[[group_col]]))[[1]][["F value"]][1]
    perm_stat
  })

  p_value <- mean(perm_stats >= observed_stat)
  list(observed_stat = observed_stat, p_value = p_value, perm_stats = perm_stats)
}

result <- permutation_test(
  data = df_test,
  group_col = "abortion_policies",
  value_col = "maternal_mortality_rates_2018_2022_per_100,000_live births"
)

cat("Observed Statistic (F):", result$observed_stat, "\n")
cat("Permutation Test P-Value:", result$p_value, "\n")
```

Null hypothesis: The maternal mortality rates do not differ across abortion policy groups (i.e., all groups come from the same distribution).

Alternate hypothesis: There is a significant difference in maternal mortality rates between the abortion policy groups.

The p-value is 0.00213, which provides provides evidence to reject the null hypothesis.

```{r}

ggplot(data.frame(perm_stats = result$perm_stats), aes(x = perm_stats)) +
  geom_histogram(binwidth = 0.1, fill = "blue", color = "black") +
  geom_vline(xintercept = result$observed_stat, color = "red", linetype = "dashed") +
  labs(
    title = "Permutation Test Distribution",
    x = "F-statistics from Permutations",
    y = "Frequency"
  ) +
  theme_minimal()
```

The observed F-statistic (red dashed line) is far to the right of this distribution, indicating its extreme nature compared to the null hypothesis.

```{r}

pairwise_permutation_test <- function(data, group_col, value_col, n_permutations = 10000) {
  groups <- unique(data[[group_col]])
  comparisons <- combn(groups, 2, simplify = FALSE)
  results <- lapply(comparisons, function(pair) {
    group1 <- data[[value_col]][data[[group_col]] == pair[1]]
    group2 <- data[[value_col]][data[[group_col]] == pair[2]]
    
    observed_diff <- abs(median(group1) - median(group2))
    
    pooled <- c(group1, group2)
    perm_diffs <- replicate(n_permutations, {
      shuffled <- sample(pooled)
      abs(median(shuffled[1:length(group1)]) - median(shuffled[(length(group1) + 1):length(pooled)]))
    })
    
    p_value <- mean(perm_diffs >= observed_diff)
    
    list(pair = pair, observed_diff = observed_diff, p_value = p_value)
  })
  
  data.frame(
    Group1 = sapply(results, function(x) x$pair[1]),
    Group2 = sapply(results, function(x) x$pair[2]),
    Observed_Diff = sapply(results, function(x) x$observed_diff),
    P_Value = sapply(results, function(x) x$p_value)
  )
}

pairwise_results <- pairwise_permutation_test(
  data = df_test,
  group_col = "abortion_policies",
  value_col = "maternal_mortality_rates_2018_2022_per_100,000_live births"
)

print(pairwise_results)
```

most restrictive & very protective observed difference = 21 p-val = 0.0105

very restrictive & protective observed difference = 21 p-val = 0.0473

```{r}

library(ggplot2)
library(tidyr)
library(dplyr)

```

```{r}

policy_order <- c("most protective", "very protective", "protective", 
                 "some restrictions/protections", "restrictive", 
                 "very restrictive", "most restrictive")
```

```{r}

create_policy_viz <- function(data, metric_col, title, y_label, plot_type = "bar") {
  processed_data <- data %>%
    filter(!is.na(abortion_policies) & 
           abortion_policies != "" & 
           abortion_policies != "NA") %>%
    mutate(abortion_policies = tolower(trimws(abortion_policies))) %>%
    group_by(abortion_policies) %>%
    summarise(avg_value = mean(!!sym(metric_col), na.rm = TRUE)) %>%
    filter(abortion_policies %in% policy_order) %>%
    mutate(
      abortion_policies = factor(abortion_policies, levels = policy_order),
      policy_numeric = as.numeric(factor(abortion_policies, levels = policy_order))
    )
  
  p <- ggplot(processed_data, aes(x = abortion_policies, y = avg_value))
  
  if(plot_type == "scatter") {
    model <- lm(avg_value ~ policy_numeric, data = processed_data)
    r_squared <- round(summary(model)$r.squared, 3)
    
    p <- p +
      geom_point(size = 4, color = "steelblue") +
      geom_smooth(aes(x = policy_numeric), method = "lm", se = TRUE, color = "red") +
      annotate("text", 
              x = length(policy_order) - 0.5,
              y = max(processed_data$avg_value),
              label = paste0("R² = ", r_squared),
              size = 4, hjust = 1)
  } else {
    p <- p + geom_bar(stat = "identity", fill = "steelblue")
  }
  
  p + labs(
    title = title,
    x = "Abortion Policy",
    y = y_label
  ) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
      axis.title = element_text(size = 12)
    )
}

```

```{r}

maternal_mortality <- create_policy_viz(df,
   "maternal_mortality_rates_2018_2022_per_100,000_live births",
   "Maternal Mortality Rate, 2018-2022",
   "Rate",
   "scatter")

```

## Women Who Went Without Care Due to Cost, by Abortion Policy

```{r}

# Kruskal-Wallis test
kruskal.test(women_18_44_who_went_without_care_because_of_cost ~ abortion_policies, data = df)

# If Kruskal-Wallis is significant, run pairwise Wilcoxon tests
pairwise.wilcox.test(df$women_18_44_who_went_without_care_because_of_cost, 
                     df$abortion_policies,
                     p.adjust.method = "bonferroni")


# Create the plot
# accompanying boxplot visualization
ggplot(df, aes(x = abortion_policies, y = women_18_44_who_went_without_care_because_of_cost, fill=abortion_policies)) +
  geom_boxplot() +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 8, angle = 45, hjust = 1),
        axis.title.y = element_text(size = 10)) + 
  scale_fill_manual(values = c("#1c7416", "#68bb59", "#acdf87", "#fab733", "#ff6242", "#ff0000", "#c61a09")) +
  labs(x = "Abortion Policies", 
       y = "% Women Without Care per 100,000",
       title = "Percentage of Women who went Without Care due to Cost by Abortion Policy") +
  theme(plot.title = element_text(size = 10, hjust = 0.5))

```

The Kruskal-Wallis test indicates a statistically significant difference in the proportion of women (ages 18-44) who went without care because of cost across at least two abortion policy groups (p=0.04237). Although the Kruskal-Wallis test suggests overall differences between groups, the pairwise comparisons reveal no statistically significant differences between specific groups

```{r}

set.seed(5100)

permutation_test <- function(data, group_col, value_col, n_permutations = 100000) {
  observed_stat <- summary(aov(data[[value_col]] ~ data[[group_col]]))[[1]][["F value"]][1]

  perm_stats <- replicate(n_permutations, {
    shuffled_values <- sample(data[[value_col]])
    perm_stat <- summary(aov(shuffled_values ~ data[[group_col]]))[[1]][["F value"]][1]
    perm_stat
  })

  p_value <- mean(perm_stats >= observed_stat)
  list(observed_stat = observed_stat, p_value = p_value, perm_stats = perm_stats)
}

result <- permutation_test(
  data = df_test,
  group_col = "abortion_policies",
  value_col = "women_18_44_who_went_without_care_because_of_cost"
)

cat("Observed Statistic (F):", result$observed_stat, "\n")
cat("Permutation Test P-Value:", result$p_value, "\n")

```

Null hypothesis: Women who went without care due to cost do not differ across abortion policy groups

Alternate hypothesis: There is a significant difference in ranking between the abortion policy groups.

The p-value is 0.0341, which provides provides overwhelming evidence to reject the null hypothesis.

```{r}

ggplot(data.frame(perm_stats = result$perm_stats), aes(x = perm_stats)) +
  geom_histogram(binwidth = 0.1, fill = "blue", color = "black") +
  geom_vline(xintercept = result$observed_stat, color = "red", linetype = "dashed") +
  labs(
    title = "Permutation Test Distribution",
    x = "F-statistics from Permutations",
    y = "Frequency"
  ) +
  theme_minimal()

```

The observed F-statistic (red dashed line) is far to the right of this distribution, indicating its extreme nature compared to the null hypothesis.

```{r}

pairwise_permutation_test <- function(data, group_col, value_col, n_permutations = 10000) {
  groups <- unique(data[[group_col]])
  comparisons <- combn(groups, 2, simplify = FALSE)
  results <- lapply(comparisons, function(pair) {
    group1 <- data[[value_col]][data[[group_col]] == pair[1]]
    group2 <- data[[value_col]][data[[group_col]] == pair[2]]
    
    observed_diff <- abs(median(group1) - median(group2))
    
    pooled <- c(group1, group2)
    perm_diffs <- replicate(n_permutations, {
      shuffled <- sample(pooled)
      abs(median(shuffled[1:length(group1)]) - median(shuffled[(length(group1) + 1):length(pooled)]))
    })
    
    p_value <- mean(perm_diffs >= observed_diff)
    
    list(pair = pair, observed_diff = observed_diff, p_value = p_value)
  })
  
  data.frame(
    Group1 = sapply(results, function(x) x$pair[1]),
    Group2 = sapply(results, function(x) x$pair[2]),
    Observed_Diff = sapply(results, function(x) x$observed_diff),
    P_Value = sapply(results, function(x) x$p_value)
  )
}

pairwise_results <- pairwise_permutation_test(
  data = df_test,
  group_col = "abortion_policies",
  value_col = "women_18_44_who_went_without_care_because_of_cost.1"
)

print(pairwise_results)
                           
```

very restrictive & very protective observed difference = 5.5 p-val = 0.0134

very restrictive & protective observed difference = 3.5 p-val = 0.0477

## Percent of Women Uninsured During Pregnancy

```{r}

# Kruskal-Wallis test
kruskal.test(`percent_uninsured_women_during_pregnancy_2021` ~ abortion_policies, data = df)

# If Kruskal-Wallis is significant, run pairwise Wilcoxon tests
pairwise.wilcox.test(df$`percent_uninsured_women_during_pregnancy_2021`, 
                     df$abortion_policies,
                     p.adjust.method = "bonferroni")


# Create the plot
# accompanying boxplot visualization
ggplot(df, aes(x = abortion_policies, y = percent_uninsured_women_during_pregnancy_2021, fill=abortion_policies)) +
  geom_boxplot() +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 8, angle = 45, hjust = 1),
        axis.title.y = element_text(size = 10)) + 
  scale_fill_manual(values = c("#1c7416", "#68bb59", "#acdf87", "#fab733", "#ff6242", "#ff0000", "#c61a09")) +
  labs(x = "Abortion Policies", 
       y = "% Women Uninsured",
       title = "Percentage of Women Uninsured During Pregnancy \nby Abortion Policy") +
  theme(plot.title = element_text(size = 10, hjust = 0.5))
```

The p-val is 0.1169, showing no significant relationship among different groups.

```{r}

set.seed(5100)

permutation_test <- function(data, group_col, value_col, n_permutations = 100000) {
  observed_stat <- summary(aov(data[[value_col]] ~ data[[group_col]]))[[1]][["F value"]][1]

  perm_stats <- replicate(n_permutations, {
    shuffled_values <- sample(data[[value_col]])
    perm_stat <- summary(aov(shuffled_values ~ data[[group_col]]))[[1]][["F value"]][1]
    perm_stat
  })

  p_value <- mean(perm_stats >= observed_stat)
  list(observed_stat = observed_stat, p_value = p_value, perm_stats = perm_stats)
}

result <- permutation_test(
  data = df_test,
  group_col = "abortion_policies",
  value_col = "percent_uninsured_women_during_pregnancy_2021"
)

cat("Observed Statistic (F):", result$observed_stat, "\n")
cat("Permutation Test P-Value:", result$p_value, "\n")

```

Null hypothesis: Women who are uninsured during pregnancy do not differ across abortion policy groups

Alternate hypothesis: There is a significant difference in ranking between the abortion policy groups.

The p-value is 0.0786

```{r}

ggplot(data.frame(perm_stats = result$perm_stats), aes(x = perm_stats)) +
  geom_histogram(binwidth = 0.1, fill = "blue", color = "black") +
  geom_vline(xintercept = result$observed_stat, color = "red", linetype = "dashed") +
  labs(
    title = "Permutation Test Distribution",
    x = "F-statistics from Permutations",
    y = "Frequency"
  ) +
  theme_minimal()

```

## Percent Uninsured Women Ages 18-64

```{r}

# Kruskal-Wallis test
kruskal.test(`percent_uninsured_women_ages 19_64` ~ abortion_policies, data = df)

# If Kruskal-Wallis is significant, run pairwise Wilcoxon tests
pairwise.wilcox.test(df$`percent_uninsured_women_ages 19_64`, 
                     df$abortion_policies,
                     p.adjust.method = "bonferroni")


# Create the plot
ggplot(df, aes(x = abortion_policies, y = `percent_uninsured_women_ages 19_64`)) +
  geom_boxplot() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "Abortion Policies", 
       y = "Percent Uninsured Women Ages 19-64",
       title = "Percent Uninsured Women Ages 19-64 by Abortion Policy")
```

P-val is 0.1044

```{r}

set.seed(5100)

permutation_test <- function(data, group_col, value_col, n_permutations = 100000) {
  observed_stat <- summary(aov(data[[value_col]] ~ data[[group_col]]))[[1]][["F value"]][1]

  perm_stats <- replicate(n_permutations, {
    shuffled_values <- sample(data[[value_col]])
    perm_stat <- summary(aov(shuffled_values ~ data[[group_col]]))[[1]][["F value"]][1]
    perm_stat
  })

  p_value <- mean(perm_stats >= observed_stat)
  list(observed_stat = observed_stat, p_value = p_value, perm_stats = perm_stats)
}

result <- permutation_test(
  data = df_test,
  group_col = "abortion_policies",
  value_col = "percent_uninsured_women_ages 19_64"
)

cat("Observed Statistic (F):", result$observed_stat, "\n")
cat("Permutation Test P-Value:", result$p_value, "\n")
```

Null hypothesis: Women who are uninsured do not differ across abortion policy groups

Alternate hypothesis: There is a significant difference in ranking between the abortion policy groups.

P-val is .12828