---
title: "Maternal Health EDA"
author: "Kristin Lloyd"
format: 
  html:
    embed-resources: true
    code-fold: true
editor: visual
---

```{r, warning=FALSE, message=FALSE}

library(tidyverse)
library(readr)
```

```{r}

df <- read_csv("data/clean_data/merged_data.csv", show_col_types = FALSE)
df <- df[!is.na(df$abortion_policies), ]
```

```{r}

table(df$abortion_policies)
```

```{r}

# Kruskal-Wallis test
kruskal.test(foster_children_per_capita ~ abortion_policies, data = df)

# If Kruskal-Wallis is significant, run pairwise Wilcoxon tests
pairwise.wilcox.test(df$foster_children_per_capita, 
                     df$abortion_policies,
                     p.adjust.method = "bonferroni")

# For visualization
# First reorder the factor levels
df$abortion_policies <- factor(df$abortion_policies, 
                             levels = c("most restrictive", 
                                      "very restrictive", 
                                      "restrictive",
                                      "some restrictions/protections",
                                      "protective",
                                      "very protective",
                                      "most protective"))

# Then create the plot
ggplot(df, aes(x = abortion_policies, y = foster_children_per_capita)) +
  geom_boxplot() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "Abortion Policies", 
       y = "Foster Children per Capita",
       title = "Foster Care Rates by Abortion Policy")
```

```{r}

set.seed(5100)

df_test <- df[df$State != "US", ]

permutation_test <- function(data, group_col, value_col, n_permutations = 100000) { observed_stat <- summary(aov(data[[value_col]] ~ data[[group_col]]))[[1]][["F value"]][1]

perm_stats <- replicate(n_permutations, { shuffled_values <- sample(data[[value_col]]) perm_stat <- summary(aov(shuffled_values ~ data[[group_col]]))[[1]][["F value"]][1] perm_stat })

p_value <- mean(perm_stats >= observed_stat) list(observed_stat = observed_stat, p_value = p_value, perm_stats = perm_stats) }

result <- permutation_test( data = df_test, group_col = "abortion_policies", value_col = "overall_health_ranking" )

cat("Observed Statistic (F):", result$observed_stat, "\n")
cat("Permutation Test P-Value:", result$p_value, "\n 

```

Null hypothesis: The health rankings do not differ across abortion policy groups (i.e., all groups come from the same distribution).

Alternate hypothesis: There is a significant difference in overall_health_ranking between the abortion policy groups.

The p-value is 0.00001, which provides provides overwhelming evidence to reject the null hypothesis.

```{r}

ggplot(data.frame(perm_stats = result$perm_stats), aes(x = perm_stats)) +
  geom_histogram(binwidth = 0.1, fill = "blue", color = "black") +
  geom_vline(xintercept = result$observed_stat, color = "red", linetype = "dashed") +
  labs(
    title = "Permutation Test Distribution",
    x = "F-statistics from Permutations",
    y = "Frequency"
  ) +
  theme_minimal()

```

The observed F-statistic (red dashed line) is far to the right of this distribution, indicating its extreme nature compared to the null hypothesis. ​

```{r}

pairwise_permutation_test <- function(data, group_col, value_col, n_permutations = 10000) {
  groups <- unique(data[[group_col]])
  comparisons <- combn(groups, 2, simplify = FALSE)
  results <- lapply(comparisons, function(pair) {
    group1 <- data[[value_col]][data[[group_col]] == pair[1]]
    group2 <- data[[value_col]][data[[group_col]] == pair[2]]
    
    observed_diff <- abs(median(group1) - median(group2))
    
    pooled <- c(group1, group2)
    perm_diffs <- replicate(n_permutations, {
      shuffled <- sample(pooled)
      abs(median(shuffled[1:length(group1)]) - median(shuffled[(length(group1) + 1):length(pooled)]))
    })
    
    p_value <- mean(perm_diffs >= observed_diff)
    
    list(pair = pair, observed_diff = observed_diff, p_value = p_value)
  })
  
  data.frame(
    Group1 = sapply(results, function(x) x$pair[1]),
    Group2 = sapply(results, function(x) x$pair[2]),
    Observed_Diff = sapply(results, function(x) x$observed_diff),
    P_Value = sapply(results, function(x) x$p_value)
  )
}

pairwise_results <- pairwise_permutation_test(
  data = df_test,
  group_col = "abortion_policies",
  value_col = "overall_health_ranking"
)

print(pairwise_results)
                           
```

most restrictive & protective observed difference = 26.5 p-val = 0.0001

```{r}

{r}

# Kruskal-Wallis test
kruskal.test(foster_children_per_capita ~ abortion_policies, data = df)

# If Kruskal-Wallis is significant, run pairwise Wilcoxon tests
pairwise.wilcox.test(df$foster_children_per_capita, 
                     df$abortion_policies,
                     p.adjust.method = "bonferroni")

# For visualization
# First reorder the factor levels
df$abortion_policies <- factor(df$abortion_policies, 
                             levels = c("most restrictive", 
                                      "very restrictive", 
                                      "restrictive",
                                      "some restrictions/protections",
                                      "protective",
                                      "very protective",
                                      "most protective"))

# Then create the plot
ggplot(df, aes(x = abortion_policies, y = foster_children_per_capita)) +
  geom_boxplot() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "Abortion Policies", 
       y = "Foster Children per Capita",
       title = "Foster Care Rates by Abortion Policy")

```

```{r}

set.seed(5100)

df_test <- df[df$State != "US", ]

permutation_test <- function(data, group_col, value_col, n_permutations = 100000) { observed_stat <- summary(aov(data[[value_col]] ~ data[[group_col]]))[[1]][["F value"]][1]

perm_stats <- replicate(n_permutations, { shuffled_values <- sample(data[[value_col]]) perm_stat <- summary(aov(shuffled_values ~ data[[group_col]]))[[1]][["F value"]][1] perm_stat })

p_value <- mean(perm_stats >= observed_stat) list(observed_stat = observed_stat, p_value = p_value, perm_stats = perm_stats) }

result <- permutation_test( data = df_test, group_col = "abortion_policies", value_col = "overall_health_ranking" )

cat("Observed Statistic (F):", result$observed_stat, "\n")
cat("Permutation Test P-Value:", result$p_value, "\n 

```

```{r}

ggplot(data.frame(perm_stats = result$perm_stats), aes(x = perm_stats)) +
  geom_histogram(binwidth = 0.1, fill = "blue", color = "black") +
  geom_vline(xintercept = result$observed_stat, color = "red", linetype = "dashed") +
  labs(
    title = "Permutation Test Distribution",
    x = "F-statistics from Permutations",
    y = "Frequency"
  ) +
  theme_minimal()

```

```{r}

pairwise_permutation_test <- function(data, group_col, value_col, n_permutations = 10000) {
  groups <- unique(data[[group_col]])
  comparisons <- combn(groups, 2, simplify = FALSE)
  results <- lapply(comparisons, function(pair) {
    group1 <- data[[value_col]][data[[group_col]] == pair[1]]
    group2 <- data[[value_col]][data[[group_col]] == pair[2]]
    
    observed_diff <- abs(median(group1) - median(group2))
    
    pooled <- c(group1, group2)
    perm_diffs <- replicate(n_permutations, {
      shuffled <- sample(pooled)
      abs(median(shuffled[1:length(group1)]) - median(shuffled[(length(group1) + 1):length(pooled)]))
    })
    
    p_value <- mean(perm_diffs >= observed_diff)
    
    list(pair = pair, observed_diff = observed_diff, p_value = p_value)
  })
  
  data.frame(
    Group1 = sapply(results, function(x) x$pair[1]),
    Group2 = sapply(results, function(x) x$pair[2]),
    Observed_Diff = sapply(results, function(x) x$observed_diff),
    P_Value = sapply(results, function(x) x$p_value)
  )
}

pairwise_results <- pairwise_permutation_test(
  data = df_test,
  group_col = "abortion_policies",
  value_col = "overall_health_ranking"
)

print(pairwise_results)

```

```{r}

# Kruskal-Wallis test
kruskal.test(foster_children_per_capita ~ abortion_policies, data = df)

# If Kruskal-Wallis is significant, run pairwise Wilcoxon tests
pairwise.wilcox.test(df$foster_children_per_capita, 
                     df$abortion_policies,
                     p.adjust.method = "bonferroni")

# For visualization
# First reorder the factor levels
df$abortion_policies <- factor(df$abortion_policies, 
                             levels = c("most restrictive", 
                                      "very restrictive", 
                                      "restrictive",
                                      "some restrictions/protections",
                                      "protective",
                                      "very protective",
                                      "most protective"))

# Then create the plot
ggplot(df, aes(x = abortion_policies, y = foster_children_per_capita)) +
  geom_boxplot() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "Abortion Policies", 
       y = "Foster Children per Capita",
       title = "Foster Care Rates by Abortion Policy")

```

```{r}

set.seed(5100)

df_test <- df[df$State != "US", ]

permutation_test <- function(data, group_col, value_col, n_permutations = 100000) { observed_stat <- summary(aov(data[[value_col]] ~ data[[group_col]]))[[1]][["F value"]][1]

perm_stats <- replicate(n_permutations, { shuffled_values <- sample(data[[value_col]]) perm_stat <- summary(aov(shuffled_values ~ data[[group_col]]))[[1]][["F value"]][1] perm_stat })

p_value <- mean(perm_stats >= observed_stat) list(observed_stat = observed_stat, p_value = p_value, perm_stats = perm_stats) }

result <- permutation_test( data = df_test, group_col = "abortion_policies", value_col = "overall_health_ranking" )

cat("Observed Statistic (F):", result$observed_stat, "\n")
cat("Permutation Test P-Value:", result$p_value, "\n 

```

```{r}

ggplot(data.frame(perm_stats = result$perm_stats), aes(x = perm_stats)) +
  geom_histogram(binwidth = 0.1, fill = "blue", color = "black") +
  geom_vline(xintercept = result$observed_stat, color = "red", linetype = "dashed") +
  labs(
    title = "Permutation Test Distribution",
    x = "F-statistics from Permutations",
    y = "Frequency"
  ) +
  theme_minimal()

```

```{r}

pairwise_permutation_test <- function(data, group_col, value_col, n_permutations = 10000) {
  groups <- unique(data[[group_col]])
  comparisons <- combn(groups, 2, simplify = FALSE)
  results <- lapply(comparisons, function(pair) {
    group1 <- data[[value_col]][data[[group_col]] == pair[1]]
    group2 <- data[[value_col]][data[[group_col]] == pair[2]]
    
    observed_diff <- abs(median(group1) - median(group2))
    
    pooled <- c(group1, group2)
    perm_diffs <- replicate(n_permutations, {
      shuffled <- sample(pooled)
      abs(median(shuffled[1:length(group1)]) - median(shuffled[(length(group1) + 1):length(pooled)]))
    })
    
    p_value <- mean(perm_diffs >= observed_diff)
    
    list(pair = pair, observed_diff = observed_diff, p_value = p_value)
  })
  
  data.frame(
    Group1 = sapply(results, function(x) x$pair[1]),
    Group2 = sapply(results, function(x) x$pair[2]),
    Observed_Diff = sapply(results, function(x) x$observed_diff),
    P_Value = sapply(results, function(x) x$p_value)
  )
}

pairwise_results <- pairwise_permutation_test(
  data = df_test,
  group_col = "abortion_policies",
  value_col = "overall_health_ranking"
)

print(pairwise_results)
                           
```