---
title: "Healthcare EDA"
author: “Kristin Lloyd”
format: 
  html:
    embed-resources: true
    code-fold: true
---

```{r}

library(readr)
df <- read_csv("../data/clean_data/merged_data.csv")

head(df)

```

```{r}

options(repos = c(CRAN = "https://cran.r-project.org"))
install.packages("dplyr")
library(dplyr)

```

```{r}

head(df)
colnames(df)

```

```{r}

library(ggplot2)
library(tidyr)
library(dplyr)

```

```{r}

# wide to long format

df_racial_mortality <- pivot_longer(
    data = df,
    cols = c(
        maternal_mortality_mothers_race_is_black_2017_2022,
        maternal_mortality_mothers_race_is_white_2017_2022
    ),
    names_to = "Race",
    values_to = "MortalityRate"
)

df_racial_mortality$Race <- ifelse(
    df_racial_mortality$Race == "maternal_mortality_mothers_race_is_black_2017_2022",
    "Black",
    ifelse(
        df_racial_mortality$Race == "maternal_mortality_mothers_race_is_white_2017_2022",
        "White",
        df_racial_mortality$Race
    )
)

```

```{r}

# box plot for maternal mortality by race

ggplot(
    data = df_racial_mortality,
    aes(x = Race, y = MortalityRate, fill = Race)
) +
    geom_boxplot(alpha = 0.7) +
    labs(
        title = "Distribution of Maternal Mortality Rates by Race (2017-2022)",
        x = "Race",
        y = "Mortality Rate per 100,000 Live Births",
        caption = "Data source: Merged healthcare dataset"
    ) +
    scale_fill_manual(
        values = c("Black" = "#1f77b4", "White" = "#ff7f0e")
    ) +
    theme_minimal() +
    theme(
        plot.title = element_text(hjust = 0.5, size = 14),
        axis.title = element_text(size = 12),
        legend.position = "none"
    )

```

###

Null hypothesis: The maternal mortality rate for Black mothers is less than or equal to the rate for White mothers.

Alternate hypothesis: The maternal mortality rate for Black mothers is greater than the rate for White mothers

```{r}

black_rates <- na.omit(df$maternal_mortality_mothers_race_is_black_2017_2022)
white_rates <- na.omit(df$maternal_mortality_mothers_race_is_white_2017_2022)

t_test_result <- t.test(black_rates, white_rates, alternative = "greater")

print(t_test_result)

```

###

The p value is extremely small and less than .05. Therefore, we can reject the null hypothesis and say that black maternal mortality rates are greater than White maternal mortality rates. The lower bound, 4.91, says that we are 95% confident the true difference in means is greater than 4.91. 

```{r}

df_long_infant <- pivot_longer(
    data = df,
    cols = c(
        total_infant_mortality_white,
        total_infant_mortality_black,
        total_infant_mortality_hispanic
    ),
    names_to = "Group",
    values_to = "InfantMortalityRate"
)

df_long_infant$Group <- ifelse(
    df_long_infant$Group == "total_infant_mortality_white", "White",
    ifelse(
        df_long_infant$Group == "total_infant_mortality_black", "Black",
        ifelse(
            df_long_infant$Group == "total_infant_mortality_hispanic", "Hispanic",
            df_long_infant$Group
        )
    )
)

```

```{r}

ggplot(
    data = df_long_infant,
    aes(x = Group, y = InfantMortalityRate, fill = Group)
) +
    geom_boxplot(alpha = 0.7) +
    labs(
        title = "Distribution of Infant Mortality Rates by Race/Ethnicity",
        x = "Race/Ethnicity",
        y = "Infant Mortality Rate per 1,000 Live Births",
        caption = "Data source: Merged healthcare dataset"
    ) +
    scale_fill_manual(
        values = c("White" = "#1f77b4", "Black" = "#ff7f0e", "Hispanic" = "#2ca02c")
    ) +
    theme_minimal() +
    theme(
        plot.title = element_text(hjust = 0.5, size = 14),
        axis.title = element_text(size = 12),
        legend.position = "none"
    )

```

###
Need to use a different statistical test for this since there are three groups with different variances. Or we can do 3 separate t-tests.

```{r}

mortality_plot <- pivot_longer(
    data = df,
    cols = starts_with("infant_mortality_"),
    names_to = "year",
    values_to = "mortality_rate"
)

mortality_plot$year <- as.numeric(substr(
    mortality_plot$year,
    nchar("infant_mortality_") + 1,
    nchar(mortality_plot$year)
))

mortality_plot <- mortality_plot[
    mortality_plot$abortion_policies == "most restrictive" &
        !is.na(mortality_plot$abortion_policies),
]

ggplot(
    data = mortality_plot,
    aes(
        x = year, y = mortality_rate,
        color = State, group = State
    )
) +
    geom_line(linewidth = 1) +
    geom_point(size = 2) +
    theme_minimal() +
    labs(
        title = "Infant Mortality Rates in Most Restrictive States (2017-2022)",
        x = "Year",
        y = "Infant Mortality Rate per 1,000 Live Births",
        color = "State",
        caption = "Data source: Merged healthcare dataset"
    ) +
    scale_x_continuous(breaks = 2017:2022) +
    theme(
        legend.position = "right",
        plot.title = element_text(hjust = 0.5, size = 14,
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10)
    )

```

###
Some go up, some go down. Not very telling!

```{r}

# Add a new column for OB/GYN proportion in the original data frame
df <- df %>%
    filter(!is.na(abortion_policies) &
        abortion_policies != "" &
        abortion_policies != "NA") %>%
    mutate(
        abortion_policies = tolower(trimws(abortion_policies)),
        obgyn_proportion = (number_of_obstetricians_and_gynecologists_employed_2023 /
            total_sumber_of_women_aged_15_44_2017) * 100000
    )

```

```{r}

create_ban_type_viz <- function(data, metric_col, title, y_label, plot_type = "bar") {
    ban_order <- c(
        "total_abortion_ban",
        "week_ban_6",
        "week_ban_12",
        "week_ban_15",
        "week_ban_18_to_23",
        "week_ban_24_to_26",
        "no_gestational_ban"
    )

    reshaped_data <- data %>%
        select(all_of(ban_order), !!sym(metric_col)) %>%
        pivot_longer(
            cols = all_of(ban_order),
            names_to = "ban_type",
            values_to = "has_ban"
        ) %>%
        filter(has_ban == 1) %>%
        group_by(ban_type) %>%
        summarise(avg_value = mean(!!sym(metric_col), na.rm = TRUE)) %>%
        mutate(
            ban_type = factor(ban_type, levels = ban_order),
            ban_numeric = as.numeric(factor(ban_type, levels = ban_order))
        )

    p <- ggplot(reshaped_data, aes(x = ban_type, y = avg_value))

    if (plot_type == "scatter") {
        model <- lm(avg_value ~ ban_numeric, data = reshaped_data)
        r_squared <- round(summary(model)$r.squared, 3)

        p <- p +
            geom_point(size = 4, color = "steelblue") +
            geom_smooth(aes(x = ban_numeric), method = "lm", se = TRUE, color = "red") +
            annotate("text",
                x = length(ban_order) - 0.5,
                y = max(reshaped_data$avg_value),
                label = paste0("R² = ", r_squared),
                size = 4, hjust = 1
            )
    } else {
        p <- p + geom_bar(stat = "identity", fill = "steelblue")
    }

    p + labs(
        title = title,
        x = "Abortion Ban Type",
        y = y_label
    ) +
        theme_minimal() +
        theme(
            axis.text.x = element_text(angle = 45, hjust = 1),
            plot.title = element_text(hjust = 0.5, size = 14),
            axis.title = element_text(size = 12)
        )
}

```

```{r}

all_cause_mortality <- create_ban_type_viz(
    df, "all_cause_mortality_rate_per_100,000_women_ages 15_44_2022",
    "All Cause Mortality Rate Women 15-44, 2022",
    "Rate"
)

```

```{r}

print(all_cause_mortality)

```

###

The total abortion ban category is significantly higher despite all of the other categories being somewhat stable. 

# Maternal Mortality

```{r}

df_filtered <- df %>%
    filter(!is.na(`maternal_mortality_rates_2018_2022_per_100,000_live births`)) %>%
    filter(State != "US") %>%
    bind_rows(tibble(
        State = "US Average",
        `maternal_mortality_rates_2018_2022_per_100,000_live births` = 23.2,
        abortion_policies = "US Average"
    )) %>%
    mutate(
        abortion_policies = factor(
            abortion_policies,
            levels = c(
                "most restrictive", "very restrictive", "restrictive",
                "some restrictions/protections",
                "protective", "very protective", "most protective", "US Average"
            )
        )
    )

ggplot(
    df_filtered,
    aes(
        y = reorder(State, `maternal_mortality_rates_2018_2022_per_100,000_live births`),
        x = `maternal_mortality_rates_2018_2022_per_100,000_live births`,
        fill = abortion_policies
    )
) +
    geom_bar(stat = "identity", width = 0.7) +
    geom_text(aes(label = sprintf("%.1f", `maternal_mortality_rates_2018_2022_per_100,000_live births`)),
        hjust = -0.2,
        size = 3
    ) +
    scale_fill_manual(values = c(
        "most restrictive" = "#bd0000",
        "very restrictive" = "#ff0000",
        "restrictive" = "#ff866b",
        "some restrictions/protections" = "#ffe857",
        "protective" = "#b6ff83",
        "very protective" = "#2ae81e",
        "most protective" = "#1e9400",
        "US Average" = "#0056df"
    )) +
    theme_minimal() +
    labs(
        title = "Maternal Mortality Rates (2018-2022) per 100,000 Live Births by State",
        x = "Maternal Mortality Rate per 100,000 Live Births",
        y = "",
        fill = "Abortion Policy"
    ) +
    theme(
        plot.title = element_text(hjust = 0.5, size = 12),
        axis.title.x = element_text(size = 10),
        axis.text.y = element_text(size = 8, color = "grey30"),
        axis.text.x = element_text(size = 8),
        legend.position = "right",
        legend.title = element_text(size = 9),
        legend.text = element_text(size = 8),
        panel.grid.major.y = element_blank(),
        panel.grid.minor = element_blank(),
        plot.background = element_rect(fill = "white", color = NA),
        panel.background = element_rect(fill = "white", color = NA)
    ) +
    scale_x_continuous(
        limits = c(0, 45),
        breaks = seq(0, 40, by = 10),
        expand = expansion(mult = c(0, 0.1))
    )

```

```{r}

df_filtered <- df %>%
    filter(!is.na(`maternal_mortality_rates_2018_2022_per_100,000_live births`)) %>%
    filter(State != "US" & State != "Ohio" & State != "Nevada") %>%
    mutate(
        simplified_policies = case_when(
            abortion_policies %in% c("most restrictive", "very restrictive", "restrictive") ~ "Restrictive",
            abortion_policies %in% c("most protective", "very protective", "protective") ~ "Protective",
            TRUE ~ abortion_policies
        )
    ) %>%
    bind_rows(tibble(
        State = "US Average",
        `maternal_mortality_rates_2018_2022_per_100,000_live births` = 23.2,
        simplified_policies = "US Average"
    ))

ggplot(
    df_filtered,
    aes(
        y = reorder(State, `maternal_mortality_rates_2018_2022_per_100,000_live births`),
        x = `maternal_mortality_rates_2018_2022_per_100,000_live births`,
        fill = simplified_policies
    )
) +
    geom_bar(stat = "identity", width = 0.7) +
    geom_text(aes(label = sprintf("%.1f", `maternal_mortality_rates_2018_2022_per_100,000_live births`)),
        hjust = -0.2,
        size = 3
    ) +
    scale_fill_manual(values = c(
        "Restrictive" = "#ff0000",
        "Protective" = "#4dbd00",
        "US Average" = "#006eff"
    )) +
    theme_minimal() +
    labs(
        title = "Maternal Mortality Rates (2018-2022) per 100,000 Live Births by State",
        x = "Maternal Mortality Rate per 100,000 Live Births",
        y = "",
        fill = "Abortion Policy"
    ) +
    theme(
        plot.title = element_text(hjust = 0.5, size = 12),
        axis.title.x = element_text(size = 10),
        axis.text.y = element_text(size = 8, color = "grey30"),
        axis.text.x = element_text(size = 8),
        legend.position = "right",
        legend.title = element_text(size = 9),
        legend.text = element_text(size = 8),
        panel.grid.major.y = element_blank(),
        panel.grid.minor = element_blank(),
        plot.background = element_rect(fill = "white", color = NA),
        panel.background = element_rect(fill = "white", color = NA)
    ) +
    scale_x_continuous(
        limits = c(0, 45),
        breaks = seq(0, 40, by = 10),
        expand = expansion(mult = c(0, 0.1))
    )

```

###
Separated only by protective and restrictive.

```{r}

install.packages("usmap")
library(usmap)

library()
abortion_map_data <- df %>%
    select(State, `percent_increase_in_clinician_provided_abortions_since_2020`) %>%
    filter(!is.na(`percent_increase_in_clinician_provided_abortions_since_2020`)) %>%
    mutate(
        state = State,
        change = `percent_increase_in_clinician_provided_abortions_since_2020`
    )

print("First few rows of our data:")
head(abortion_map_data)

capitals_df <- data.frame(
  long = c(-116.200, -100.346, -93.603, -92.189, -92.289,
          -91.187, -97.743, -97.503, -90.182, -86.279,
          -84.280, -86.781, -84.878, -86.162, -81.633),
  lat = c(43.617, 44.367, 41.591, 38.579, 34.746,
          30.457, 30.266, 35.482, 32.298, 32.377,
          30.438, 36.162, 38.186, 39.768, 38.336),
  state = c("Idaho", "South Dakota", "Iowa", "Missouri", "Arkansas",
            "Louisiana", "Texas", "Oklahoma", "Mississippi", "Alabama",
            "Florida", "Tennessee", "Kentucky", "Indiana", "West Virginia")
)

capitals_mapped <- usmap_transform(capitals_df, input_names = c("long", "lat"))

plot_usmap(data = abortion_map_data, values = "change") + geom_point(data = capitals_mapped, 
             aes(x = x, y = y),
             shape = 8,        # Star shape
             size = 4,         # Size of stars
             color = "black",  # Color of stars
             stroke = 1.5) +   # Thickness of stars
    scale_fill_gradient2(
        name = "% Change",
        low = "#ff6b6b",
        mid = "#ffffff",
        high = "#4dbd00",
        midpoint = 0,
        na.value = "gray80"
    ) +
    theme(
        legend.position = "right",
        plot.title = element_text(size = 14, hjust = 0.5)
    ) +
    labs(title = "Change in Clinician-Provided Abortions Since 2020 by State", caption = "* indicate Most Restrictive States")

```

###
Cool, you can see it goes up near the abortion ban states.

```{r}

library(corrplot)

healthcare_metrics <- df %>%
    select(
        overall_health_ranking,
        coverage_access_and_affordability,
        health_care_quality_and_prevention,
        health_outcomes,
        women_18_44_who_went_without_care_because_of_cost.1
    ) %>%
    rename(
        "Overall Health Ranking" = overall_health_ranking,
        "Coverage & Access" = coverage_access_and_affordability,
        "Healthcare Quality" = health_care_quality_and_prevention,
        "Health Outcomes" = health_outcomes,
        "Care Cost Barriers" = women_18_44_who_went_without_care_because_of_cost.1
    )

cor_matrix <- cor(healthcare_metrics, use = "complete.obs")

corrplot(cor_matrix,
    method = "color",
    type = "upper",
    tl.col = "black",
    tl.srt = 45,
    addCoef.col = "black",
    number.cex = 0.7,
    col = colorRampPalette(c("#E41A1C", "#FFFFFF", "#4DAF4A"))(200),
    tl.cex = 0.7,
    cl.cex = 0.7,
    diag = FALSE,
    mar = c(0, 0, 1, 0),
    title = "Healthcare Metrics Correlations",
    family = "Arial",
    bg = "white",
    addgrid.col = "gray90"
)

title("Correlations Between Healthcare Metrics",
    line = -0.5,
    cex.main = 1.2,
    font.main = 2
)

```

###
Love this

```{r}

library(corrplot)

healthcare_metrics <- df %>%
    select(
        overall_health_ranking,
        coverage_access_and_affordability,
        health_care_quality_and_prevention,
        health_outcomes,
        women_18_44_who_went_without_care_because_of_cost.1,
        `percent_uninsured_women_ages 19_64`,
        `teen_birth_rate_per_1,000_population_ages_15_19_2022.1`,
        `maternal_mortality_rates_2018_2022_per_100,000_live births`
    ) %>%
    rename(
        "Overall Health Ranking" = overall_health_ranking,
        "Coverage & Access" = coverage_access_and_affordability,
        "Healthcare Quality" = health_care_quality_and_prevention,
        "Health Outcomes" = health_outcomes,
        "Care Cost Barriers" = women_18_44_who_went_without_care_because_of_cost.1,
        "Uninsured Women %" = `percent_uninsured_women_ages 19_64`,
        "Teen Birth Rate" = `teen_birth_rate_per_1,000_population_ages_15_19_2022.1`,
        "Maternal Mortality" = `maternal_mortality_rates_2018_2022_per_100,000_live births`
    )

cor_matrix <- cor(healthcare_metrics, use = "complete.obs")

corrplot(cor_matrix,
    method = "color",
    type = "upper",
    tl.col = "black",
    tl.srt = 45,
    addCoef.col = "black",
    number.cex = 0.7,
    col = colorRampPalette(c("#E41A1C", "#FFFFFF", "#4DAF4A"))(200),
    tl.cex = 0.7,
    cl.cex = 0.7,
    diag = FALSE,
    mar = c(0, 0, 1, 0),
    title = "Healthcare Metrics Correlations",
    family = "Arial",
    bg = "white",
    addgrid.col = "gray90"
)

title("Correlations Between Healthcare Metrics",
    line = -0.5,
    cex.main = 1.2,
    font.main = 2
)

```

###
Can add more, have to decide which.

```{r}

# Convert abortion policies to a numeric scale (fix potential missing levels issue)
data$abortion_policies_numeric <- as.numeric(factor(data$abortion_policies,
    levels = c(
        "most restrictive", "very restrictive",
        "restrictive", "some restrictions/protections",
        "protective", "very protective", "most protective"
    )
))

# Subset the relevant variables (ensure consistent column naming)
relevant_data <- data[, c(
    "abortion_policies_numeric",
    "twenty_five_and_older_had_bachelors_by_higher_2021",
    "teen_birth_rate_per_1,000_population_ages_15_19_2022",
    "maternity_care_workforce_per_100,000_women_ages_15_44"
)]

# Rename columns (ensure the number of new names matches the columns)
colnames(relevant_data) <- c("abortion_policies", "education", "teen_birth_rate", "maternity_care_workforce")

# Remove rows with missing values
relevant_data <- na.omit(relevant_data)

# Calculate correlation matrix
correlation_matrix <- cor(relevant_data, method = "spearman")

# Print correlation matrix
print(correlation_matrix)


```

```{r}

install.packages("corrplot")
library(corrplot)

corrplot(correlation_matrix, method = "circle", type = "lower", tl.col = "black", tl.srt = 45)

```

### Interpretation

Teen Birth Rate and Education (-0.8109):

Strong negative correlation: As teen birth rates increase, education levels decrease significantly. This is a strong inverse relationship.

Teen Birth Rate and Abortion Policies (-0.6947):

Moderate negative correlation: States with more protective abortion policies tend to have lower teen birth rates.

Education and Abortion Policies (+0.7318):

Strong positive correlation: States with more protective abortion policies also tend to have higher levels of education.

Teen Birth Rate and Maternity Care Workforce (-0.5486):

Moderate negative correlation: Higher maternity care workforce density correlates with lower teen birth rates.

Abortion Policies and Maternity Care Workforce (0.7326):

Strong positive correlation: States with more protective abortion policies tend to have a denser maternity care workforce.
Education and Maternity Care Workforce (0.6766):

Moderate positive correlation: Higher levels of education correlate with better maternity care workforce density.

Can we say, "Teen births disrupt educational attainment and career development, leading to a cycle where mothers are less likely to enter or remain in the workforce".

More teen births equal less education and less women in the workforce. Strict abortion policies have more teen births. This leads to a cycle.

```{r}

# Install and load necessary packages
install.packages("ggrepel")
library(ggrepel)
library(ggplot2)

# Create the scatter plot with custom cluster colors
ggplot(health_metrics, aes(x = health_outcomes, y = overall_health_ranking, color = cluster)) +
    geom_point(size = 3) +
    geom_text_repel(aes(label = abortion_policies), size = 3) +
    scale_color_manual(values = c("1" = "#008e17", "2" = "#cf0000", "3" = "#fad100")) +
    labs(
        title = "Clusters of States by Health Metrics and Abortion Policies",
        x = "Health Outcomes",
        y = "Overall Health Ranking"
    ) +
    theme_minimal()

```

### Interpretation

Kind of cool because looks at two variables and you can clearly see higher rankings are restrictive states.


```{r}
# Boxplot by abortion policies
ggplot(df, aes(x = abortion_policies, y = women_18_44_who_went_without_care_because_of_cost.1, fill = abortion_policies)) +
    geom_boxplot() +
    scale_fill_manual(values = policy_colors) +
    labs(
        title = "Women Without Care Due to Cost Across Abortion Policies",
        x = "Abortion Policies",
        y = "Percentage of Women Without Care"
    ) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

### 

Medians of restrictive groups are higher! Bigger variability with most restriction but even the low is kind of high compared to protective and most protective.

