---
title: "Healthcare Permutation Tests"
author: “Kristin Lloyd”
format: 
  html:
    embed-resources: true
    code-fold: true
---

```{r}

library(readr)
df <- read_csv("../data/clean_data/merged_data.csv")

head(df)

```

```{r}

options(repos = c(CRAN = "https://cran.r-project.org"))
install.packages("dplyr")
library(dplyr)

```

```{r}

library(ggplot2)
library(tidyr)
library(dplyr)

```

# Permutation Tests

## Overall Health Ranking

```{r}

set.seed(5100)

df_test <- df[df$State != "US", ]

set.seed(5100)

permutation_test <- function(data, group_col, value_col, n_permutations = 100000) {
  observed_stat <- summary(aov(data[[value_col]] ~ data[[group_col]]))[[1]][["F value"]][1]

  perm_stats <- replicate(n_permutations, {
    shuffled_values <- sample(data[[value_col]])
    perm_stat <- summary(aov(shuffled_values ~ data[[group_col]]))[[1]][["F value"]][1]
    perm_stat
  })

  p_value <- mean(perm_stats >= observed_stat)
  list(observed_stat = observed_stat, p_value = p_value, perm_stats = perm_stats)
}

result <- permutation_test(
  data = df_test,
  group_col = "abortion_policies",
  value_col = "overall_health_ranking"
)

cat("Observed Statistic (F):", result$observed_stat, "\n")
cat("Permutation Test P-Value:", result$p_value, "\n")

```

### Interpretation

Null hypothesis: The health rankings do not differ across abortion policy groups (i.e., all groups come from the same distribution).

Alternate hypothesis: There is a significant difference in overall_health_ranking between the abortion policy groups.

The p-value is 0.00001, which provides provides overwhelming evidence to reject the null hypothesis.

```{r}

ggplot(data.frame(perm_stats = result$perm_stats), aes(x = perm_stats)) +
  geom_histogram(binwidth = 0.1, fill = "blue", color = "black") +
  geom_vline(xintercept = result$observed_stat, color = "red", linetype = "dashed") +
  labs(
    title = "Permutation Test Distribution",
    x = "F-statistics from Permutations",
    y = "Frequency"
  ) +
  theme_minimal()

```

### Interpretation

The observed F-statistic (red dashed line) is far to the right of this distribution, indicating its extreme nature compared to the null hypothesis.
​
```{r}

pairwise_permutation_test <- function(data, group_col, value_col, n_permutations = 10000) {
  groups <- unique(data[[group_col]])
  comparisons <- combn(groups, 2, simplify = FALSE)
  results <- lapply(comparisons, function(pair) {
    group1 <- data[[value_col]][data[[group_col]] == pair[1]]
    group2 <- data[[value_col]][data[[group_col]] == pair[2]]
    
    observed_diff <- abs(median(group1) - median(group2))
    
    pooled <- c(group1, group2)
    perm_diffs <- replicate(n_permutations, {
      shuffled <- sample(pooled)
      abs(median(shuffled[1:length(group1)]) - median(shuffled[(length(group1) + 1):length(pooled)]))
    })
    
    p_value <- mean(perm_diffs >= observed_diff)
    
    list(pair = pair, observed_diff = observed_diff, p_value = p_value)
  })
  
  data.frame(
    Group1 = sapply(results, function(x) x$pair[1]),
    Group2 = sapply(results, function(x) x$pair[2]),
    Observed_Diff = sapply(results, function(x) x$observed_diff),
    P_Value = sapply(results, function(x) x$p_value)
  )
}

pairwise_results <- pairwise_permutation_test(
  data = df_test,
  group_col = "abortion_policies",
  value_col = "overall_health_ranking"
)

print(pairwise_results)

```

### Interpretation

most restrictive & protective
observed difference = 26.5
p-val =	0.0001

most restrictive & very protective
observed difference = 20.5
p-val = 0.0139

most restrictive & some restrictions/protections
observed difference = 14.5
p-val = 0.0396

most restrictive & restrictive
observed difference = 13.5
p-val = 0.0150

most restrictive & most protective
observed difference = 31.0
p-val = 0.0064

protective & very restrictive
observed difference = 24.0
p-val = 0.0091

## Coverage, Access, Affordability

```{r}

set.seed(5100)

permutation_test <- function(data, group_col, value_col, n_permutations = 100000) {
  observed_stat <- summary(aov(data[[value_col]] ~ data[[group_col]]))[[1]][["F value"]][1]

  perm_stats <- replicate(n_permutations, {
    shuffled_values <- sample(data[[value_col]])
    perm_stat <- summary(aov(shuffled_values ~ data[[group_col]]))[[1]][["F value"]][1]
    perm_stat
  })

  p_value <- mean(perm_stats >= observed_stat)
  list(observed_stat = observed_stat, p_value = p_value, perm_stats = perm_stats)
}

result <- permutation_test(
  data = df_test,
  group_col = "abortion_policies",
  value_col = "coverage_access_affordability"
)

cat("Observed Statistic (F):", result$observed_stat, "\n")
cat("Permutation Test P-Value:", result$p_value, "\n")

```

### Interpretation

Null hypothesis: The coverage, access, and affordability rankings do not differ across abortion policy groups

Alternate hypothesis: There is a significant difference in ranking between the abortion policy groups.

The p-value is 0.00001, which provides provides overwhelming evidence to reject the null hypothesis.


```{r}

ggplot(data.frame(perm_stats = result$perm_stats), aes(x = perm_stats)) +
  geom_histogram(binwidth = 0.1, fill = "blue", color = "black") +
  geom_vline(xintercept = result$observed_stat, color = "red", linetype = "dashed") +
  labs(
    title = "Permutation Test Distribution",
    x = "F-statistics from Permutations",
    y = "Frequency"
  ) +
  theme_minimal()

```

### Interpretation

The observed F-statistic (red dashed line) is far to the right of this distribution, indicating its extreme nature compared to the null hypothesis.

```{r}

pairwise_permutation_test <- function(data, group_col, value_col, n_permutations = 10000) {
  groups <- unique(data[[group_col]])
  comparisons <- combn(groups, 2, simplify = FALSE)
  results <- lapply(comparisons, function(pair) {
    group1 <- data[[value_col]][data[[group_col]] == pair[1]]
    group2 <- data[[value_col]][data[[group_col]] == pair[2]]
    
    observed_diff <- abs(median(group1) - median(group2))
    
    pooled <- c(group1, group2)
    perm_diffs <- replicate(n_permutations, {
      shuffled <- sample(pooled)
      abs(median(shuffled[1:length(group1)]) - median(shuffled[(length(group1) + 1):length(pooled)]))
    })
    
    p_value <- mean(perm_diffs >= observed_diff)
    
    list(pair = pair, observed_diff = observed_diff, p_value = p_value)
  })
  
  data.frame(
    Group1 = sapply(results, function(x) x$pair[1]),
    Group2 = sapply(results, function(x) x$pair[2]),
    Observed_Diff = sapply(results, function(x) x$observed_diff),
    P_Value = sapply(results, function(x) x$p_value)
  )
}

pairwise_results <- pairwise_permutation_test(
  data = df_test,
  group_col = "abortion_policies",
  value_col = "coverage_access_affordability"
)

print(pairwise_results)

```

### Interpretation

###
most restrictive & protective
observed difference = 26.5
p-val =	0.0001

most restrictive & very protective
observed difference = 20.5
p-val = 0.0139

most restrictive & some restrictions/protections
observed difference = 14.5
p-val = 0.0396

most restrictive & restrictive
observed difference = 13.5
p-val = 0.0150

most restrictive & most protective
observed difference = 31.0
p-val = 0.0064

protective & very restrictive
observed difference = 24.0
p-val = 0.0091

## Healthcare Quality Prevention

```{r}

set.seed(5100)

permutation_test <- function(data, group_col, value_col, n_permutations = 100000) {
  observed_stat <- summary(aov(data[[value_col]] ~ data[[group_col]]))[[1]][["F value"]][1]

  perm_stats <- replicate(n_permutations, {
    shuffled_values <- sample(data[[value_col]])
    perm_stat <- summary(aov(shuffled_values ~ data[[group_col]]))[[1]][["F value"]][1]
    perm_stat
  })

  p_value <- mean(perm_stats >= observed_stat)
  list(observed_stat = observed_stat, p_value = p_value, perm_stats = perm_stats)
}

result <- permutation_test(
  data = df_test,
  group_col = "abortion_policies",
  value_col = "health_care_quality_and_prevention"
)

cat("Observed Statistic (F):", result$observed_stat, "\n")
cat("Permutation Test P-Value:", result$p_value, "\n")

```

### Interpretation

Null hypothesis: The healthcare quality and prevention rankings do not differ across abortion policy groups

Alternate hypothesis: There is a significant difference in ranking between the abortion policy groups.

The p-value is 0.01934, which provides provides overwhelming evidence to reject the null hypothesis.

```{r}

ggplot(data.frame(perm_stats = result$perm_stats), aes(x = perm_stats)) +
  geom_histogram(binwidth = 0.1, fill = "blue", color = "black") +
  geom_vline(xintercept = result$observed_stat, color = "red", linetype = "dashed") +
  labs(
    title = "Permutation Test Distribution",
    x = "F-statistics from Permutations",
    y = "Frequency"
  ) +
  theme_minimal()

```

### Interpretation

The observed F-statistic (red dashed line) is far to the right of this distribution, indicating its extreme nature compared to the null hypothesis.

```{r}

pairwise_permutation_test <- function(data, group_col, value_col, n_permutations = 10000) {
  groups <- unique(data[[group_col]])
  comparisons <- combn(groups, 2, simplify = FALSE)
  results <- lapply(comparisons, function(pair) {
    group1 <- data[[value_col]][data[[group_col]] == pair[1]]
    group2 <- data[[value_col]][data[[group_col]] == pair[2]]
    
    observed_diff <- abs(median(group1) - median(group2))
    
    pooled <- c(group1, group2)
    perm_diffs <- replicate(n_permutations, {
      shuffled <- sample(pooled)
      abs(median(shuffled[1:length(group1)]) - median(shuffled[(length(group1) + 1):length(pooled)]))
    })
    
    p_value <- mean(perm_diffs >= observed_diff)
    
    list(pair = pair, observed_diff = observed_diff, p_value = p_value)
  })
  
  data.frame(
    Group1 = sapply(results, function(x) x$pair[1]),
    Group2 = sapply(results, function(x) x$pair[2]),
    Observed_Diff = sapply(results, function(x) x$observed_diff),
    P_Value = sapply(results, function(x) x$p_value)
  )
}

pairwise_results <- pairwise_permutation_test(
  data = df_test,
  group_col = "abortion_policies",
  value_col = "health_care_quality_and_prevention"
)

print(pairwise_results)

```

### Interpretation

most restrictive & protective
observed difference = 19
p-val =	0.0130

most restrictive & restrictive
observed difference = 18.5
p-val = 0.0399

most restrictive & most protective
observed difference = 26.5
p-val = 0.0117

## Health Outcomes

```{r}

set.seed(5100)

permutation_test <- function(data, group_col, value_col, n_permutations = 100000) {
  observed_stat <- summary(aov(data[[value_col]] ~ data[[group_col]]))[[1]][["F value"]][1]

  perm_stats <- replicate(n_permutations, {
    shuffled_values <- sample(data[[value_col]])
    perm_stat <- summary(aov(shuffled_values ~ data[[group_col]]))[[1]][["F value"]][1]
    perm_stat
  })

  p_value <- mean(perm_stats >= observed_stat)
  list(observed_stat = observed_stat, p_value = p_value, perm_stats = perm_stats)
}

result <- permutation_test(
  data = df_test,
  group_col = "abortion_policies",
  value_col = "health_outcomes"
)

cat("Observed Statistic (F):", result$observed_stat, "\n")
cat("Permutation Test P-Value:", result$p_value, "\n")

```

### Interpretation

Null hypothesis: The health outcomes rankings do not differ across abortion policy groups

Alternate hypothesis: There is a significant difference in ranking between the abortion policy groups.

The p-value is 0.0002, which provides provides overwhelming evidence to reject the null hypothesis.

```{r}

ggplot(data.frame(perm_stats = result$perm_stats), aes(x = perm_stats)) +
  geom_histogram(binwidth = 0.1, fill = "blue", color = "black") +
  geom_vline(xintercept = result$observed_stat, color = "red", linetype = "dashed") +
  labs(
    title = "Permutation Test Distribution",
    x = "F-statistics from Permutations",
    y = "Frequency"
  ) +
  theme_minimal()

```

### Interpretation

The observed F-statistic (red dashed line) is far to the right of this distribution, indicating its extreme nature compared to the null hypothesis.

```{r}

pairwise_permutation_test <- function(data, group_col, value_col, n_permutations = 10000) {
  groups <- unique(data[[group_col]])
  comparisons <- combn(groups, 2, simplify = FALSE)
  results <- lapply(comparisons, function(pair) {
    group1 <- data[[value_col]][data[[group_col]] == pair[1]]
    group2 <- data[[value_col]][data[[group_col]] == pair[2]]
    
    observed_diff <- abs(median(group1) - median(group2))
    
    pooled <- c(group1, group2)
    perm_diffs <- replicate(n_permutations, {
      shuffled <- sample(pooled)
      abs(median(shuffled[1:length(group1)]) - median(shuffled[(length(group1) + 1):length(pooled)]))
    })
    
    p_value <- mean(perm_diffs >= observed_diff)
    
    list(pair = pair, observed_diff = observed_diff, p_value = p_value)
  })
  
  data.frame(
    Group1 = sapply(results, function(x) x$pair[1]),
    Group2 = sapply(results, function(x) x$pair[2]),
    Observed_Diff = sapply(results, function(x) x$observed_diff),
    P_Value = sapply(results, function(x) x$p_value)
  )
}

pairwise_results <- pairwise_permutation_test(
  data = df_test,
  group_col = "abortion_policies",
  value_col = "health_outcomes"
)

print(pairwise_results)

```

### Interpretation

most restrictive & protective
observed difference = 28.0
p-val = 0.0013

most restrictive & very restrictive          
observed difference = 17.0
p-val = 0.0518

most restrictive & very protective          
observed difference = 32.5
p-val = 0.0005

most restrictive & some restrictions/protections
observed difference = 12.5
p-val = 0.0671

most restrictive & restrictive
observed difference = 20.5
p-val = 0.0062

most restrictive & most protective
observed difference = 32.0
p-val = 0.0177

## Women who went without care because of cost

```{r}

set.seed(5100)

permutation_test <- function(data, group_col, value_col, n_permutations = 100000) {
  observed_stat <- summary(aov(data[[value_col]] ~ data[[group_col]]))[[1]][["F value"]][1]

  perm_stats <- replicate(n_permutations, {
    shuffled_values <- sample(data[[value_col]])
    perm_stat <- summary(aov(shuffled_values ~ data[[group_col]]))[[1]][["F value"]][1]
    perm_stat
  })

  p_value <- mean(perm_stats >= observed_stat)
  list(observed_stat = observed_stat, p_value = p_value, perm_stats = perm_stats)
}

result <- permutation_test(
  data = df_test,
  group_col = "abortion_policies",
  value_col = "women_18_44_who_went_without_care_because_of_cost.1"
)

cat("Observed Statistic (F):", result$observed_stat, "\n")
cat("Permutation Test P-Value:", result$p_value, "\n")

```

### Interpretation

Null hypothesis: Women who went without care due to cost do not differ across abortion policy groups

Alternate hypothesis: There is a significant difference in ranking between the abortion policy groups.

The p-value is 0.04805, which provides provides overwhelming evidence to reject the null hypothesis.

```{r}

ggplot(data.frame(perm_stats = result$perm_stats), aes(x = perm_stats)) +
  geom_histogram(binwidth = 0.1, fill = "blue", color = "black") +
  geom_vline(xintercept = result$observed_stat, color = "red", linetype = "dashed") +
  labs(
    title = "Permutation Test Distribution",
    x = "F-statistics from Permutations",
    y = "Frequency"
  ) +
  theme_minimal()

```

### Interpretation

The observed F-statistic (red dashed line) is far to the right of this distribution, indicating its extreme nature compared to the null hypothesis.

```{r}

pairwise_permutation_test <- function(data, group_col, value_col, n_permutations = 10000) {
  groups <- unique(data[[group_col]])
  comparisons <- combn(groups, 2, simplify = FALSE)
  results <- lapply(comparisons, function(pair) {
    group1 <- data[[value_col]][data[[group_col]] == pair[1]]
    group2 <- data[[value_col]][data[[group_col]] == pair[2]]
    
    observed_diff <- abs(median(group1) - median(group2))
    
    pooled <- c(group1, group2)
    perm_diffs <- replicate(n_permutations, {
      shuffled <- sample(pooled)
      abs(median(shuffled[1:length(group1)]) - median(shuffled[(length(group1) + 1):length(pooled)]))
    })
    
    p_value <- mean(perm_diffs >= observed_diff)
    
    list(pair = pair, observed_diff = observed_diff, p_value = p_value)
  })
  
  data.frame(
    Group1 = sapply(results, function(x) x$pair[1]),
    Group2 = sapply(results, function(x) x$pair[2]),
    Observed_Diff = sapply(results, function(x) x$observed_diff),
    P_Value = sapply(results, function(x) x$p_value)
  )
}

pairwise_results <- pairwise_permutation_test(
  data = df_test,
  group_col = "abortion_policies",
  value_col = "women_18_44_who_went_without_care_because_of_cost.1"
)

print(pairwise_results)

```

### Interpretation

most restrictive & protective
observed difference = 4.0
p-val = 0.0719

most restrictive & very protective          
observed difference = 5.5
p-val = 0.0874

protective & very restrictive
observed difference = 4.0
p-val = 0.0473

very restrictive & very protective
observed difference = 5.5
p-val = 0.0115

very restrictive & most protective
observed difference = 6.5
p-val = 0.0671

## Uninsured Women

```{r}

set.seed(5100)

permutation_test <- function(data, group_col, value_col, n_permutations = 100000) {
  observed_stat <- summary(aov(data[[value_col]] ~ data[[group_col]]))[[1]][["F value"]][1]

  perm_stats <- replicate(n_permutations, {
    shuffled_values <- sample(data[[value_col]])
    perm_stat <- summary(aov(shuffled_values ~ data[[group_col]]))[[1]][["F value"]][1]
    perm_stat
  })

  p_value <- mean(perm_stats >= observed_stat)
  list(observed_stat = observed_stat, p_value = p_value, perm_stats = perm_stats)
}

result <- permutation_test(
  data = df_test,
  group_col = "abortion_policies",
  value_col = "percent_uninsured_women_ages 19_64"
)

cat("Observed Statistic (F):", result$observed_stat, "\n")
cat("Permutation Test P-Value:", result$p_value, "\n")

```

### Interpretation

Null hypothesis: Women who are uninsured do not differ across abortion policy groups

Alternate hypothesis: There is a significant difference in ranking between the abortion policy groups.

The p-value is 0.0335, which provides provides overwhelming evidence to reject the null hypothesis.

```{r}

ggplot(data.frame(perm_stats = result$perm_stats), aes(x = perm_stats)) +
  geom_histogram(binwidth = 0.1, fill = "blue", color = "black") +
  geom_vline(xintercept = result$observed_stat, color = "red", linetype = "dashed") +
  labs(
    title = "Permutation Test Distribution",
    x = "F-statistics from Permutations",
    y = "Frequency"
  ) +
  theme_minimal()

```

### Interpretation

The observed F-statistic (red dashed line) is far to the right of this distribution, indicating its extreme nature compared to the null hypothesis.

```{r}

pairwise_permutation_test <- function(data, group_col, value_col, n_permutations = 10000) {
  groups <- unique(data[[group_col]])
  comparisons <- combn(groups, 2, simplify = FALSE)
  results <- lapply(comparisons, function(pair) {
    group1 <- data[[value_col]][data[[group_col]] == pair[1]]
    group2 <- data[[value_col]][data[[group_col]] == pair[2]]
    
    observed_diff <- abs(median(group1) - median(group2))
    
    pooled <- c(group1, group2)
    perm_diffs <- replicate(n_permutations, {
      shuffled <- sample(pooled)
      abs(median(shuffled[1:length(group1)]) - median(shuffled[(length(group1) + 1):length(pooled)]))
    })
    
    p_value <- mean(perm_diffs >= observed_diff)
    
    list(pair = pair, observed_diff = observed_diff, p_value = p_value)
  })
  
  data.frame(
    Group1 = sapply(results, function(x) x$pair[1]),
    Group2 = sapply(results, function(x) x$pair[2]),
    Observed_Diff = sapply(results, function(x) x$observed_diff),
    P_Value = sapply(results, function(x) x$p_value)
  )
}

pairwise_results <- pairwise_permutation_test(
  data = df_test,
  group_col = "abortion_policies",
  value_col = "percent_uninsured_women_ages 19_64"
)

print(pairwise_results)

```

### Interpretation

very restrictive & very protective
observed difference = 4.0
p-val = 0.0445

## Uninsured during pregnancy

```{r}

set.seed(5100)

permutation_test <- function(data, group_col, value_col, n_permutations = 100000) {
  observed_stat <- summary(aov(data[[value_col]] ~ data[[group_col]]))[[1]][["F value"]][1]

  perm_stats <- replicate(n_permutations, {
    shuffled_values <- sample(data[[value_col]])
    perm_stat <- summary(aov(shuffled_values ~ data[[group_col]]))[[1]][["F value"]][1]
    perm_stat
  })

  p_value <- mean(perm_stats >= observed_stat)
  list(observed_stat = observed_stat, p_value = p_value, perm_stats = perm_stats)
}

result <- permutation_test(
  data = df_test,
  group_col = "abortion_policies",
  value_col = "percent_uninsured_women_during_pregnancy_2021"
)

cat("Observed Statistic (F):", result$observed_stat, "\n")
cat("Permutation Test P-Value:", result$p_value, "\n")

```

### Interpretation

Null hypothesis: Women who are uninsured during pregnancy do not differ across abortion policy groups

Alternate hypothesis: There is a significant difference in ranking between the abortion policy groups.

The p-value is 0.02205, which provides provides overwhelming evidence to reject the null hypothesis.

```{r}

ggplot(data.frame(perm_stats = result$perm_stats), aes(x = perm_stats)) +
  geom_histogram(binwidth = 0.1, fill = "blue", color = "black") +
  geom_vline(xintercept = result$observed_stat, color = "red", linetype = "dashed") +
  labs(
    title = "Permutation Test Distribution",
    x = "F-statistics from Permutations",
    y = "Frequency"
  ) +
  theme_minimal()

```

### Interpretation

The observed F-statistic (red dashed line) is far to the right of this distribution, indicating its extreme nature compared to the null hypothesis.

PAIR TEST NOT WORKING - WHY?

### Interpretation


## Teen Birth

```{r}

set.seed(510)

permutation_test <- function(data, group_col, value_col, n_permutations = 100000) {
  observed_stat <- summary(aov(data[[value_col]] ~ data[[group_col]]))[[1]][["F value"]][1]

  perm_stats <- replicate(n_permutations, {
    shuffled_values <- sample(data[[value_col]])
    perm_stat <- summary(aov(shuffled_values ~ data[[group_col]]))[[1]][["F value"]][1]
    perm_stat
  })

  p_value <- mean(perm_stats >= observed_stat)
  list(observed_stat = observed_stat, p_value = p_value, perm_stats = perm_stats)
}

result <- permutation_test(
  data = df_test,
  group_col = "abortion_policies",
  value_col = "teen_birth_rate_per_1,000_population_ages_15_19_2022.1"
)

cat("Observed Statistic (F):", result$observed_stat, "\n")
cat("Permutation Test P-Value:", result$p_value, "\n")

```

### Interpretation

Null hypothesis: Teen pregnancies do not differ across abortion policy groups

Alternate hypothesis: There is a significant difference in ranking between the abortion policy groups.

The p-value is 0.00005, which provides provides overwhelming evidence to reject the null hypothesis.

```{r}

ggplot(data.frame(perm_stats = result$perm_stats), aes(x = perm_stats)) +
  geom_histogram(binwidth = 0.1, fill = "blue", color = "black") +
  geom_vline(xintercept = result$observed_stat, color = "red", linetype = "dashed") +
  labs(
    title = "Permutation Test Distribution",
    x = "F-statistics from Permutations",
    y = "Frequency"
  ) +
  theme_minimal()

```

### Interpretation

The observed F-statistic (red dashed line) is far to the right of this distribution, indicating its extreme nature compared to the null hypothesis.

```{r}

pairwise_permutation_test <- function(data, group_col, value_col, n_permutations = 10000) {
  groups <- unique(data[[group_col]])
  comparisons <- combn(groups, 2, simplify = FALSE)
  results <- lapply(comparisons, function(pair) {
    group1 <- data[[value_col]][data[[group_col]] == pair[1]]
    group2 <- data[[value_col]][data[[group_col]] == pair[2]]
    
    observed_diff <- abs(median(group1) - median(group2))
    
    pooled <- c(group1, group2)
    perm_diffs <- replicate(n_permutations, {
      shuffled <- sample(pooled)
      abs(median(shuffled[1:length(group1)]) - median(shuffled[(length(group1) + 1):length(pooled)]))
    })
    
    p_value <- mean(perm_diffs >= observed_diff)
    
    list(pair = pair, observed_diff = observed_diff, p_value = p_value)
  })
  
  data.frame(
    Group1 = sapply(results, function(x) x$pair[1]),
    Group2 = sapply(results, function(x) x$pair[2]),
    Observed_Diff = sapply(results, function(x) x$observed_diff),
    P_Value = sapply(results, function(x) x$p_value)
  )
}

pairwise_results <- pairwise_permutation_test(
  data = df_test,
  group_col = "abortion_policies",
  value_col = "teen_birth_rate_per_1,000_population_ages_15_19_2022.1"
)

print(pairwise_results)

```

### Interpretation

most restrictive & protective
observed difference = 8.90
p-val = 0.0006

most restrictive & very restrictive 
observed difference = 5.50
p-val = 0.0683

most restrictive & very protective
observed difference = 10.30
p-val = 0.0004

most restrictive & some restrictions/predictions
observed difference = 6.10
p-val = 0.0410

most restrictive & restrictive
observed difference = 8.40
p-val = 0.0020

most restrictive & most protective
observed difference = 12.15
p-val = 0.0063

## Workforce

```{r}

set.seed(510)

permutation_test <- function(data, group_col, value_col, n_permutations = 100000) {
  observed_stat <- summary(aov(data[[value_col]] ~ data[[group_col]]))[[1]][["F value"]][1]

  perm_stats <- replicate(n_permutations, {
    shuffled_values <- sample(data[[value_col]])
    perm_stat <- summary(aov(shuffled_values ~ data[[group_col]]))[[1]][["F value"]][1]
    perm_stat
  })

  p_value <- mean(perm_stats >= observed_stat)
  list(observed_stat = observed_stat, p_value = p_value, perm_stats = perm_stats)
}

result <- permutation_test(
  data = df_test,
  group_col = "abortion_policies",
  value_col = "maternity_care_workforce_per_100,000_women_ages_15_44"
)

cat("Observed Statistic (F):", result$observed_stat, "\n")
cat("Permutation Test P-Value:", result$p_value, "\n")

```

### Interpretation

Null hypothesis: Maternal workforce does not differ across abortion policy groups

Alternate hypothesis: There is a significant difference in ranking between the abortion policy groups.

The p-value is 0.00005, which provides provides overwhelming evidence to reject the null hypothesis.

```{r}

ggplot(data.frame(perm_stats = result$perm_stats), aes(x = perm_stats)) +
  geom_histogram(binwidth = 0.1, fill = "blue", color = "black") +
  geom_vline(xintercept = result$observed_stat, color = "red", linetype = "dashed") +
  labs(
    title = "Permutation Test Distribution",
    x = "F-statistics from Permutations",
    y = "Frequency"
  ) +
  theme_minimal()

```

### Interpretation

The observed F-statistic (red dashed line) is far to the right of this distribution, indicating its extreme nature compared to the null hypothesis.

```{r}

pairwise_permutation_test <- function(data, group_col, value_col, n_permutations = 10000) {
  groups <- unique(data[[group_col]])
  comparisons <- combn(groups, 2, simplify = FALSE)
  results <- lapply(comparisons, function(pair) {
    group1 <- data[[value_col]][data[[group_col]] == pair[1]]
    group2 <- data[[value_col]][data[[group_col]] == pair[2]]
    
    observed_diff <- abs(median(group1) - median(group2))
    
    pooled <- c(group1, group2)
    perm_diffs <- replicate(n_permutations, {
      shuffled <- sample(pooled)
      abs(median(shuffled[1:length(group1)]) - median(shuffled[(length(group1) + 1):length(pooled)]))
    })
    
    p_value <- mean(perm_diffs >= observed_diff)
    
    list(pair = pair, observed_diff = observed_diff, p_value = p_value)
  })
  
  data.frame(
    Group1 = sapply(results, function(x) x$pair[1]),
    Group2 = sapply(results, function(x) x$pair[2]),
    Observed_Diff = sapply(results, function(x) x$observed_diff),
    P_Value = sapply(results, function(x) x$p_value)
  )
}

pairwise_results <- pairwise_permutation_test(
  data = df_test,
  group_col = "abortion_policies",
  value_col = "maternity_care_workforce_per_100,000_women_ages_15_44"
)

print(pairwise_results)

```

### Interpretation

most restrictive & protective 
observed difference = 38.95
p-val = 0.0002

most restrictive & very protective 
observed difference = 26.95
p-val = 0.0005

most restrictive & some restrictions/protections
observed difference = 17.75
p-val = 0.0350

most restrictive & most protective
observed difference = 52.85
p-val = 0.0063

protective & very restrictive 
observed difference = 38
p-val = 0.0225

protective & restrictive
observed difference = 31
p-val = 0.0204

very restrictive & very protective
observed_difference = 26
p-val = 0.0269

very restrictive & most protective
observed difference = 51.90
p-val = 0.0672

very protective & restriction
observed difference = 19
p-val = 0.0601

very protective & most protective
observed difference = 25.90
p-val = 0.0567

restrictive & most protective 
observed difference = 44.90
p-val = 0.0289

## Depression Screenings

```{r}

set.seed(510)

permutation_test <- function(data, group_col, value_col, n_permutations = 100000) {
  observed_stat <- summary(aov(data[[value_col]] ~ data[[group_col]]))[[1]][["F value"]][1]

  perm_stats <- replicate(n_permutations, {
    shuffled_values <- sample(data[[value_col]])
    perm_stat <- summary(aov(shuffled_values ~ data[[group_col]]))[[1]][["F value"]][1]
    perm_stat
  })

  p_value <- mean(perm_stats >= observed_stat)
  list(observed_stat = observed_stat, p_value = p_value, perm_stats = perm_stats)
}

result <- permutation_test(
  data = df_test,
  group_col = "abortion_policies",
  value_col = "percent_women_who_received_a_postpartum_depression_screening_2021"
)

cat("Observed Statistic (F):", result$observed_stat, "\n")
cat("Permutation Test P-Value:", result$p_value, "\n")

```

### Interpretation

Null hypothesis: % women who recieved a postpartum depression screening does not differ across abortion policy groups

Alternate hypothesis: There is a significant difference in ranking between the abortion policy groups.

The p-value is 0.00879, which provides provides overwhelming evidence to reject the null hypothesis.

```{r}

ggplot(data.frame(perm_stats = result$perm_stats), aes(x = perm_stats)) +
  geom_histogram(binwidth = 0.1, fill = "blue", color = "black") +
  geom_vline(xintercept = result$observed_stat, color = "red", linetype = "dashed") +
  labs(
    title = "Permutation Test Distribution",
    x = "F-statistics from Permutations",
    y = "Frequency"
  ) +
  theme_minimal()

```

### Interpretation

The observed F-statistic (red dashed line) is far to the right of this distribution, indicating its extreme nature compared to the null hypothesis.

```{r}

pairwise_permutation_test <- function(data, group_col, value_col, n_permutations = 10000) {
  groups <- unique(data[[group_col]])
  comparisons <- combn(groups, 2, simplify = FALSE)
  results <- lapply(comparisons, function(pair) {
    group1 <- data[[value_col]][data[[group_col]] == pair[1]]
    group2 <- data[[value_col]][data[[group_col]] == pair[2]]
    
    observed_diff <- abs(median(group1) - median(group2))
    
    pooled <- c(group1, group2)
    perm_diffs <- replicate(n_permutations, {
      shuffled <- sample(pooled)
      abs(median(shuffled[1:length(group1)]) - median(shuffled[(length(group1) + 1):length(pooled)]))
    })
    
    p_value <- mean(perm_diffs >= observed_diff)
    
    list(pair = pair, observed_diff = observed_diff, p_value = p_value)
  })
  
  data.frame(
    Group1 = sapply(results, function(x) x$pair[1]),
    Group2 = sapply(results, function(x) x$pair[2]),
    Observed_Diff = sapply(results, function(x) x$observed_diff),
    P_Value = sapply(results, function(x) x$p_value)
  )
}

pairwise_results <- pairwise_permutation_test(
  data = df_test,
  group_col = "abortion_policies",
  value_col = "percent_women_who_received_a_postpartum_depression_screening_2021"
)

print(pairwise_results)

```

NOT WORKING - WHY?

## 

```{r}

# percent_women_experiencing_intimate_partner_violence_2021

```

### Interpretation

### Interpretation

### Interpretation

## 

```{r}

# maternal_mortality_rates_2018_2022_per_100,000_live births

```

### Interpretation

### Interpretation

### Interpretation

## 

```{r}

# make number of obgyn proportional to number of women (both column names are in merged data), then run on that

```

### Interpretation

### Interpretation

### Interpretation

## 

```{r}

# all_cause_mortality_rate_per_100,000_women_ages 15_44_2022

```

### Interpretation

### Interpretation

### Interpretation

## 

```{r}

# 

```

### Interpretation

### Interpretation

### Interpretation

## 

```{r}

# 

```

### Interpretation

### Interpretation

### Interpretation

## 

```{r}

# 

```

### Interpretation

### Interpretation

### Interpretation

## 

```{r}

# 

```

### Interpretation

### Interpretation

### Interpretation

